---
title: "Univariate Descriptive Statistics"
---


## Learning outcomes {.unnumbered}

By the end of this chapter the reader should be able to:

- Distinguish qualitative from quantitative variables, and within each group identify the appropriate sub-type (nominal/ordinal, discrete/continuous) and level of measurement.
- Construct a complete frequency table — absolute, relative, cumulative absolute, and cumulative relative frequencies — for ungrouped and grouped data.
- Choose and produce the appropriate graphical representation (bar chart, histogram with density correction, pie chart, box-plot) for a given variable type.
- Compute and interpret the main measures of central tendency: arithmetic mean $\bar{x}$, median $Me$, mode $Mo$, and the geometric mean $G$.
- Compute and interpret measures of dispersion (range, IQR, variance $S^2$, standard deviation $S$, coefficient of variation $CV$) and apply the linear-transformation rules.
- Compute and interpret quartiles, percentiles, Fisher's skewness $g_1$ and kurtosis $g_2$.
- Build a Lorenz curve and compute the Gini index $G$ to quantify concentration in an economic distribution.

## Motivating empirical question {.unnumbered}

> *What is the typical nightly price of an AirBnB apartment in Granada, and how unequally is that price distributed across listings?*

The running example throughout this chapter is a simulated sample of 80 AirBnB nightly prices in Granada that mimics the right-skewed shape of a real short-term-rental market: most listings cluster around a moderate price, while a few luxury apartments stretch the upper tail. Quantitative Techniques I is a *descriptive* and *probabilistic* course — we summarise data, we do not yet test hypotheses about a wider population (that comes later in the curriculum). The univariate tools introduced here will be combined with bivariate methods in [Chapter 2](02-bivariate.qmd) and reinterpreted probabilistically from [Chapter 4](04-random-variables.qmd) onwards.

## What statistics is — and is not

`\lettrine[lines=3]{S}{tatistics}`{=latex} is the science of collecting, organising, analysing, interpreting, and presenting data to support effective decision-making. It is conventionally divided into two branches:

- **Descriptive statistics** — methods to organise, summarise, and present data, reducing large amounts of information into a few key numbers or graphs. The headline "the unemployment rate in Spain was $11.8\%$ in Q4 2024" is a descriptive statistic.
- **Inferential statistics** — uses a *sample* to make estimates, predictions, or decisions about a *population*. "Based on a survey of 1,500 consumers, $62\%$ prefer online shopping" is an inferential statement.

TC1 focuses on **descriptive statistics and probability**; inferential techniques (confidence intervals, hypothesis tests, sampling distributions) are deliberately left to TC2 and Econometrics I.

::: {.callout-note}
## Definition: population and sample

The **population** is the complete collection of all individuals, objects, or measurements of interest. A **sample** is a subset of the population selected for study.
:::

We work with samples rather than entire populations because full enumeration is typically too costly, too time-consuming, sometimes destructive (testing the lifespan of light bulbs requires burning them out), and sometimes simply infeasible (the population may be infinite or inaccessible). A well-chosen sample, drawn with an appropriate sampling method, can represent the population accurately.

## Statistical variables {#sec-variables}

A **statistical variable** is a characteristic that can take different values across the individuals in a population. Variables are classified along two crossing dimensions: the *kind of value* they can take, and the *level of measurement* they support.

### Types of variables

- **Qualitative (categorical)** variables express qualities or attributes.
  - **Nominal**: categories with no natural order — eye colour, nationality, religion.
  - **Ordinal**: categories with a meaningful order, but with non-quantifiable differences between them — education level (primary / secondary / university), satisfaction (low / medium / high).
- **Quantitative (numerical)** variables take numerical values and admit arithmetic operations.
  - **Discrete**: only isolated values, typically integers — number of children, number of employees.
  - **Continuous**: any value within an interval — height, weight, income, temperature.

### Levels of measurement

There are four levels of measurement, ordered from least to most informative:

1. **Nominal** — categories with no natural order. Allowed operations: $=$, $\neq$. Example: eye colour.
2. **Ordinal** — ordered categories with non-quantifiable differences. Allowed: $=$, $\neq$, $<$, $>$. Example: education level.
3. **Interval** — ordered, equal differences are meaningful, but there is no true zero. Allowed: $+$, $-$. Example: temperature in $^\circ$C (since $0\,^\circ$C does not mean "no temperature").
4. **Ratio** — like interval but with a meaningful zero. All arithmetic operations apply. Example: income (€), distance (km), weight (kg). Statements like "twice as much" are meaningful.

::: {.callout-note}
## Example: classifying variables

| Variable | Type | Level |
|---|---|---|
| Gender | Qualitative | Nominal |
| Customer satisfaction (1–5) | Qualitative | Ordinal |
| Number of employees | Quantitative, discrete | Ratio |
| Monthly income (€) | Quantitative, continuous | Ratio |
| Temperature ($^\circ$C) | Quantitative, continuous | Interval |
:::

::: {.callout-warning}
## Common mistake: averaging nominal data

The type of variable dictates which statistics are meaningful. The "average eye colour" is nonsensical, but the average income is not. Always check the level of measurement before applying an arithmetic summary.
:::

## Frequency tables and graphs

### Ungrouped data {#sec-freqtable}

When a variable takes a manageable number of distinct values, the data are organised in a **frequency table**.

::: {.callout-note}
## Definition: frequency-table notation

Let $x_1, x_2, \ldots, x_k$ be the $k$ distinct values taken by the variable $X$ over $n$ observations.

- $n_i$: **absolute frequency** — how many times value $x_i$ appears.
- $f_i = n_i / n$: **relative frequency** — the proportion of observations equal to $x_i$.
- $N_i = \sum_{j=1}^{i} n_j$: **cumulative absolute frequency**.
- $F_i = N_i / n$: **cumulative relative frequency**.
:::

The defining identities are
$$
\sum_{i=1}^{k} n_i = n, \qquad \sum_{i=1}^{k} f_i = 1, \qquad N_k = n, \qquad F_k = 1.
$$

::: {.callout-note}
## Example: number of bedrooms in 70 houses

| $x_i$ | $n_i$ | $f_i$ | $N_i$ | $F_i$ |
|:---:|:---:|:---:|:---:|:---:|
| 1 | 7 | 0.10 | 7 | 0.10 |
| 2 | 14 | 0.20 | 21 | 0.30 |
| 3 | 21 | 0.30 | 42 | 0.60 |
| 4 | 21 | 0.30 | 63 | 0.90 |
| 5 | 7 | 0.10 | 70 | 1.00 |

Sixty per cent of houses have three or fewer bedrooms (read from $F_3 = 0.60$). Both 3 and 4 bedrooms appear with the same maximal frequency, so the distribution is *bimodal*.
:::

### Grouped data (intervals)

When a continuous variable takes too many distinct values to tabulate one-by-one, observations are grouped into **class intervals**, conventionally left-closed and right-open: $[L_i, L_{i+1})$.

::: {.callout-note}
## Definition: grouped-data notation

For data grouped into intervals $[L_i, L_{i+1})$:

- $c_i = (L_i + L_{i+1})/2$: the **class mark** (midpoint), used as a representative value.
- $a_i = L_{i+1} - L_i$: the **class width** (amplitude).
- $h_i = n_i / a_i$: the **frequency density**, used in histograms when intervals have unequal widths.
:::

::: {.callout-note}
## Example: hourly wages of 100 workers

| Interval | $c_i$ | $n_i$ | $a_i$ | $h_i$ | $f_i$ |
|:---:|:---:|:---:|:---:|:---:|:---:|
| $[0, 10)$ | 5 | 25 | 10 | 2.50 | 0.25 |
| $[10, 20)$ | 15 | 40 | 10 | 4.00 | 0.40 |
| $[20, 40)$ | 30 | 20 | 20 | 1.00 | 0.20 |
| $[40, 50)$ | 45 | 15 | 10 | 1.50 | 0.15 |

The interval $[20, 40)$ has width $20$, double the others. Using the frequency $n_i$ as the height of a histogram bar over that interval would visually exaggerate its weight; using the density $h_i = n_i / a_i$ makes the *area* of each bar proportional to the frequency, which is the correct visual encoding.
:::

### Graphical representations

- **Bar charts** are for qualitative and discrete quantitative variables. Each bar's height equals its frequency, and bars are separated by gaps.
- **Histograms** are for continuous variables. Bars touch — no gaps. When intervals have unequal widths, the height of each bar must be the *density* $h_i = n_i / a_i$, not the frequency.
- **Pie charts** are admissible for any variable type. Each sector's angle equals $f_i \times 360^\circ$. They are misleading with many categories, in which case a bar chart is preferable.
- **Box-plots** summarise five statistics in one figure — discussed in [Section 1.5.4](#sec-boxplot).

::: {.callout-warning}
## Common mistake: truncated axes

Always check the scale of the vertical axis before interpreting a graph. A vertical axis that starts above zero can make tiny differences look like landslides — a classic technique in misleading political and corporate graphics.
:::

## Measures of central tendency {#sec-central}

Measures of central tendency identify a "typical" or "representative" value for the data.

### Moments

Before specific summaries, it is helpful to introduce **moments**, which unify many statistics in a single framework. (Summation notation is reviewed in [Appendix A](appendix-a-prerequisites.qmd).)

::: {.callout-note}
## Definition: moments

The $r$-th **non-centred moment** (about the origin) is
$$
a_r = \frac{1}{n}\sum_{i=1}^{k} x_i^r \, n_i = \sum_{i=1}^{k} x_i^r \, f_i.
$$
The $r$-th **centred moment** (about the mean) is
$$
m_r = \frac{1}{n}\sum_{i=1}^{k} (x_i - \bar{x})^r \, n_i.
$$
:::

Special cases are $a_1 = \bar{x}$, $m_1 = 0$, $m_2 = S^2$ (variance), $m_3$ enters skewness, and $m_4$ enters kurtosis.

### Arithmetic mean

::: {.callout-note}
## Definition: arithmetic mean

The arithmetic mean of a sample with distinct values $x_1, \ldots, x_k$ and frequencies $n_1, \ldots, n_k$ is
:::

$$
\bar{x} = \frac{1}{n}\sum_{i=1}^{k} x_i\, n_i = \sum_{i=1}^{k} x_i\, f_i.
$$

For grouped data the class mark $c_i$ replaces $x_i$.

**Properties.**

1. All observations contribute to its calculation.
2. It is **sensitive to outliers**: extreme values pull the mean toward them.
3. The sum of deviations from the mean is zero: $\sum (x_i - \bar{x})\, n_i = 0$.
4. **Linear transformation**: if $Y = a + bX$, then $\bar{y} = a + b\bar{x}$.

::: {.callout-note}
## Example: pocket money

Weekly pocket money (€) for 13 children:
$$
5,\, 5,\, 5,\, 5,\, 5,\, 5,\, 6,\, 6,\, 6,\, 6,\, 6,\, 30,\, 40.
$$
$$
\bar{x} = \frac{6(5) + 5(6) + 30 + 40}{13} = \frac{130}{13} = 10\,\text{€}.
$$

The mean is €10, yet 11 of the 13 children actually receive €5 or €6. The two extreme observations pull the mean upward — a textbook case where the mean is not representative.
:::

::: {.callout-note}
## Example: linear transformation — parking revenue

The average parking time at a Granada parking lot is $\bar{x} = 37$ minutes. The fee is $Y = 0.30 + 0.015 X$ euros, where $X$ is the time in minutes. The average revenue per vehicle is
$$
\bar{y} = 0.30 + 0.015 \times 37 = 0.855\,\text{€}.
$$
No individual parking times are needed: the linear-transformation property delivers $\bar{y}$ from $\bar{x}$ alone.
:::

### Geometric mean

::: {.callout-note}
## Definition: geometric mean

For a strictly positive sample,
:::

$$
G = \sqrt[n]{x_1 \cdot x_2 \cdots x_n} = \left(\prod_{i=1}^n x_i\right)^{1/n}.
$$

The geometric mean is the right tool for averaging **cumulative growth rates** (investment returns, population growth).

::: {.callout-note}
## Example: investment returns

An investor puts €10,000 into a fund. Over three years the returns are $+20\%$, $-10\%$, $+15\%$, giving growth factors $1.20,\, 0.90,\, 1.15$.

$$
G = \sqrt[3]{1.20 \times 0.90 \times 1.15} = \sqrt[3]{1.242} = 1.0748.
$$

Average annual return: $7.48\%$. The *arithmetic* mean of the returns, $(20 - 10 + 15)/3 = 8.33\%$, would **overestimate** the true average growth.
:::

### Mode

The **mode** $Mo$ is the value of the variable with the highest frequency. A distribution can be unimodal, bimodal, or multimodal. For grouped data with *equal* widths, the modal class is the one with the largest $n_i$ and $Mo$ is approximated by its midpoint; with *unequal* widths, compare *densities* $h_i$ instead. The mode can be computed for any level of measurement (including nominal data), is insensitive to outliers, but may fail to be unique or even to exist.

### Median

::: {.callout-note}
## Definition: median

The median $Me$ is the value that divides the ordered distribution into two equal halves: $50\%$ of observations below it, $50\%$ above.
:::

For **ungrouped** data, sort in ascending order and take
$$
Me = \begin{cases} x_{(n+1)/2} & \text{if $n$ is odd,} \\ \dfrac{x_{n/2} + x_{n/2+1}}{2} & \text{if $n$ is even.} \end{cases}
$$

For **grouped** data, locate the *median interval* — the first one with $N_i \geq n/2$ — and use the linear-interpolation formula
$$
Me = L_i + \frac{n/2 - N_{i-1}}{n_i}\, a_i,
$$
where $L_i$ is the lower bound of the median interval, $N_{i-1}$ is the cumulative frequency *before* it, and $a_i$ is the interval width.

The median is **robust to outliers** and is the preferred summary for skewed distributions (income, house prices, AirBnB nightly rates).

::: {.callout-note}
## Example: mean vs median for the pocket-money sample

With $n = 13$ (odd), the median is $Me = x_{7} = 6$ €. The mean was €10; the median is €6. The median is clearly the more representative summary of the *typical* child's pocket money — the two outliers (€30, €40) cannot move it.

This is why economists routinely report median household income rather than mean income: income distributions are right-skewed, and the mean is pulled up by a small number of very high earners.
:::

::: {.callout-note}
## Example: median from grouped data — salaries

For ten workers grouped as $[0,10),\, [10,20),\, [20,30),\, [30,40)$ with frequencies $1, 2, 3, 4$:

$n/2 = 5$. Cumulative frequencies: $N_1 = 1$, $N_2 = 3$, $N_3 = 6$, $N_4 = 10$. The median interval is $[20,30)$ (the first with $N_i \geq 5$).
$$
Me = 20 + \frac{5 - 3}{3}\times 10 = 20 + 6.67 = 26.67.
$$
The median salary is approximately €2,667.
:::

### Percentiles and quartiles {#sec-quartiles}

::: {.callout-note}
## Definition: percentile

The $k$-th percentile $P_k$ is the value below which $k\%$ of the observations fall. Special cases are the **quartiles** $Q_1 = P_{25}$, $Q_2 = P_{50} = Me$, $Q_3 = P_{75}$.
:::

For grouped data the same interpolation idea gives
$$
P_k = L_i + \frac{k\,n/100 - N_{i-1}}{n_i}\, a_i,
$$
where $[L_i, L_{i+1})$ is the interval containing the percentile.

::: {.callout-note}
## Example: quartiles from the wages table

Using the hourly-wage table ($n = 100$) with cumulative frequencies $N_1=25$, $N_2=65$, $N_3=85$, $N_4=100$:

- $Q_1 = P_{25}$ lies in the first interval $[0,10)$ since $N_1 = 25 \geq 25$:
$$
Q_1 = 0 + \frac{25 - 0}{25}\times 10 = 10.
$$
- $Q_3 = P_{75}$ lies in $[20,40)$ since $N_2 = 65 < 75 \leq 85 = N_3$:
$$
Q_3 = 20 + \frac{75 - 65}{20}\times 20 = 30.
$$

The interquartile range is $IQR = Q_3 - Q_1 = 20$ €.
:::

## Measures of dispersion {#sec-dispersion}

Two datasets can share the same mean yet differ radically in spread. Consider two insurance companies, each with two clients: Company A has ages $40, 40$ and Company B has $20, 60$. Both have mean $40$, but the mean is "representative" only in the first case. Dispersion measures quantify this idea.

### Range and interquartile range

The **range** is
$$
R = x_{\max} - x_{\min}.
$$
Simple to compute, but depends only on the two extreme values and is therefore very sensitive to outliers. The **interquartile range**
$$
IQR = Q_3 - Q_1 = P_{75} - P_{25}
$$
covers the central $50\%$ of the data and is robust.

### Variance and standard deviation

::: {.callout-note}
## Definition: variance and standard deviation

The (sample) variance and standard deviation, using the divisor $n$ convention, are
:::

$$
S^2 = \frac{1}{n}\sum_{i=1}^{k} (x_i - \bar{x})^2\, n_i = a_2 - \bar{x}^2, \qquad S = \sqrt{S^2}.
$$

The second formula — "mean of squares minus square of the mean" — is usually quicker to compute by hand. The standard deviation has the same units as the variable.

::: {.callout-warning}
## Common mistake: confusing $S^2$ with the divisor-$n-1$ estimator

This book follows the Spanish business-statistics convention and uses divisor $n$. Many English textbooks divide by $n-1$ and call the result $\hat{\sigma}^2$. Both are valid; they differ by a factor of $n/(n-1)$, which is negligible for $n \geq 30$. R's `var()` uses $n-1$, so when you need the divisor-$n$ version in the lab you must multiply by $(n-1)/n$ or recompute from scratch.
:::

**Properties.**

1. $S^2 \geq 0$ always, with $S^2 = 0$ if and only if all observations equal the mean.
2. **Change of origin**: if $Y = X + a$, then $S_Y^2 = S_X^2$. Adding a constant does not change the spread.
3. **Change of scale**: if $Y = bX$, then $S_Y^2 = b^2 S_X^2$ and $S_Y = |b|\, S_X$.
4. **General linear transformation**: if $Y = a + bX$, then $S_Y^2 = b^2 S_X^2$.

::: {.callout-note}
## Example: variance of hours worked

Three-, four-, and three-hour columns at $x_i = 5, 6, 7$:
$\bar{x} = 60/10 = 6$, $a_2 = 366/10 = 36.6$, hence
$$
S^2 = 36.6 - 36 = 0.6, \qquad S = \sqrt{0.6} \approx 0.775.
$$
:::

### Coefficient of variation

::: {.callout-note}
## Definition: Pearson's coefficient of variation
$CV = S / \bar{x}$.
:::

The CV is **dimensionless**, allowing comparison of dispersion across variables with different units or scales. A smaller $CV$ means less relative dispersion and a more representative mean. The CV is invariant to changes of *scale* (it survives a re-currency conversion) but **not** to changes of *origin*. It should not be used when $\bar{x} \approx 0$.

::: {.callout-note}
## Example: salaries vs hours worked

- Salaries: $\bar{x} = 25$, $S = 10$, hence $CV = 10/25 = 0.40$.
- Hours: $\bar{x} = 6$, $S = 0.775$, hence $CV = 0.775/6 = 0.129$.

The salary distribution shows considerably more *relative* dispersion than hours worked, even though the salary variance ($100$) is much larger than the hours variance ($0.6$) — the comparison is meaningful only through the CV because the variables are on different scales.
:::

### Box-plot {#sec-boxplot}

The **box-plot** (box-and-whisker plot) is a visual summary of five statistics: the minimum (excluding outliers), $Q_1$, the median, $Q_3$, and the maximum (excluding outliers). Points beyond $Q_1 - 1.5\,IQR$ or $Q_3 + 1.5\,IQR$ are flagged as **outliers**. Box-plots are especially useful for **comparing distributions** side by side, e.g. salary distributions across departments.

## Measures of shape

### Skewness

Skewness quantifies the **asymmetry** of a distribution.

::: {.callout-note}
## Definition: Fisher's skewness coefficient

$$
g_1 = \frac{m_3}{S^3}, \qquad m_3 = \frac{1}{n}\sum_{i=1}^{k}(x_i - \bar{x})^3\, n_i.
$$
:::

**Interpretation.** $g_1 > 0$ indicates **right (positive) skewness** — the tail extends to the right; the bulk of the data sits on the left (income is the classical example). $g_1 < 0$ indicates **left (negative) skewness** — the tail extends to the left (age at retirement). $g_1 = 0$ is a *necessary* but not sufficient condition for symmetry.

The coefficient is dimensionless and invariant under linear transformations with $b > 0$.

::: {.callout-note}
## Example: skewness for the hours-worked data

With three observations at $5$, four at $6$, three at $7$:
$$
m_3 = \frac{1}{10}\big[3(-1)^3 + 4(0)^3 + 3(1)^3\big] = 0.
$$
Hence $g_1 = 0$: the distribution is perfectly symmetric.
:::

There is also a quick-and-dirty **Pearson** skewness coefficient $A_p = 3(\bar{x} - Me)/S$, used informally when only $\bar{x}$, $Me$, and $S$ are available.

### Kurtosis

Kurtosis quantifies the "peakedness" and **tail heaviness** of a distribution relative to a normal benchmark.

::: {.callout-note}
## Definition: Fisher's kurtosis coefficient

$$
g_2 = \frac{m_4}{S^4} - 3, \qquad m_4 = \frac{1}{n}\sum_{i=1}^{k}(x_i - \bar{x})^4\, n_i.
$$
:::

The subtraction of $3$ makes the normal distribution the reference point. $g_2 < 0$ is **platykurtic** (flatter than normal, light tails); $g_2 = 0$ is **mesokurtic** (normal-like); $g_2 > 0$ is **leptokurtic** (sharper peak, heavy tails, more outlier-prone).

::: {.callout-note}
## Aside: fat tails in finance

Stock returns are typically leptokurtic. Extreme events — crashes and booms — happen more often than a normal model predicts. This is why simple Gaussian models underestimate financial risk, and why kurtosis matters in market-risk management.
:::

## Measures of concentration {#sec-concentration}

**Concentration** measures the degree of *inequality* in how a variable's total is distributed across individuals. Two extreme cases bookend the spectrum:

- **Perfect equality (equidistribution)**: every individual receives the same amount.
- **Maximum concentration**: one individual receives everything; the rest get nothing.

Applications are pervasive in economics: income and wealth inequality, market concentration, land ownership, tax-burden distribution.

### The Lorenz curve

::: {.callout-note}
## Definition: Lorenz curve

The Lorenz curve graphs concentration. Sort the data from smallest to largest, then plot, for each $i$,

- $p_i$: the cumulative percentage of *population*,
- $q_i$: the cumulative percentage of the *variable's total*.
:::

The curve always passes through $(0, 0)$ and $(1, 1)$. The $45^\circ$ diagonal $q = p$ is the line of perfect equality. The further the Lorenz curve bows below the diagonal, the greater the concentration. The shaded area between the curve and the diagonal drives the Gini index.

### Gini index

::: {.callout-note}
## Definition: Gini index

$$G = \frac{\text{Area between Lorenz curve and diagonal}}{\text{Area of the triangle below the diagonal}}.$$

A practical computational formula, with $p_0 = q_0 = 0$, is
:::

$$
G = 1 - \sum_{i=1}^{k}(p_i - p_{i-1})(q_i + q_{i-1}).
$$

**Interpretation.** $G = 0$ is perfect equality; $G = 1$ is maximum concentration. Real-world country income Ginis run roughly from $0.25$ (Scandinavia) to $0.65$ (South Africa). Spain's INE 2023 figure is about $0.327$.

::: {.callout-note}
## Example: Gini for ten salaries

Salaries (hundreds of €) grouped at midpoints $5, 15, 25, 35$ with frequencies $1, 2, 3, 4$. Total income $= 250$. The cumulative shares are $p_i = 0.10, 0.30, 0.60, 1.00$ and $q_i = 0.02, 0.14, 0.44, 1.00$.

\begin{align*}
G &= 1 - \big[(0.10)(0.02) + (0.20)(0.16) + (0.30)(0.58) + (0.40)(1.44)\big] \\
  &= 1 - [0.002 + 0.032 + 0.174 + 0.576] = 1 - 0.784 = 0.216.
\end{align*}

A Gini of about $0.22$ indicates relatively low salary inequality in this small sample.
:::

### The mediala (briefly)

The **mediala** is the value that splits the distribution so that the sum of all values *below* it equals the sum of all values *above* it. When concentration is weak, the mediala is close to the median. When concentration is strong (few individuals account for most of the total), the mediala lies well above the median.

## R Lab — AirBnB nightly prices in Granada

The lab uses a simulated sample of 80 nightly prices designed to look like a typical short-term-rental market: a positively skewed bulk plus a few luxury listings in the upper tail. The chapter-wide seed is `set.seed(2026)`, but to remain consistent with the LearnR tutorial that uses the same dataset we keep the lab seed at $42$.

```{r ch01-setup}
#| message: false
#| warning: false
# Base R is enough for everything in this lab; we add nothing exotic.
```

### Generating the data

```{r ch01-data}
set.seed(42)
prices <- round(c(rlnorm(75, meanlog = log(70), sdlog = 0.45),
                  runif(5, 250, 450)), 0)

length(prices)
head(prices, 10)
summary(prices)
```

The `summary()` output already hints at right skewness: the mean exceeds the median, and the maximum sits well beyond $Q_3$.

### Frequency table

We use `cut()` to bin prices into 40-euro-wide intervals and then build $(n_i, f_i, N_i, F_i)$.

```{r ch01-freq-table}
breaks   <- seq(20, 460, by = 40)
classes  <- cut(prices, breaks = breaks, right = FALSE)
ni       <- table(classes)
fi       <- prop.table(ni)
Ni       <- cumsum(ni)
Fi       <- cumsum(fi)

freq_table <- data.frame(
  Interval = names(ni),
  ni = as.integer(ni),
  fi = round(as.numeric(fi), 3),
  Ni = as.integer(Ni),
  Fi = round(as.numeric(Fi), 3)
)
knitr::kable(freq_table,
             caption = "Grouped frequency table of AirBnB nightly prices")
```

The first two or three classes (20–100 €) hold most apartments. The right tail is sparse — the visual signature of positive skewness.

### Central tendency

```{r ch01-central}
stat_mode <- function(x) {
  # Returns the most frequent value (the first one, in case of ties).
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}

c(mean   = mean(prices),
  median = median(prices),
  mode   = stat_mode(prices),
  geo    = exp(mean(log(prices))))
```

Mean above median is the textbook signature of right skewness. The geometric mean lies below the arithmetic mean, as the AM–GM inequality guarantees for non-constant positive data.

### Effect of an outlier

```{r ch01-outlier}
prices_out <- c(prices, 5000)  # a single luxury penthouse

c(mean_without   = mean(prices),
  mean_with      = mean(prices_out),
  median_without = median(prices),
  median_with    = median(prices_out))
```

A single €5,000 listing moves the mean substantially but barely nudges the median. This is the canonical illustration of why the median is the robust default for skewed economic variables.

### Dispersion and the coefficient of variation

R's `var()` and `sd()` divide by $n-1$. To recover the divisor-$n$ statistics defined in the chapter we rescale.

```{r ch01-dispersion}
n      <- length(prices)
S2     <- sum((prices - mean(prices))^2) / n    # divisor n
S      <- sqrt(S2)
CV_pct <- S / mean(prices) * 100

c(variance_n = round(S2, 2),
  sd_n       = round(S, 2),
  IQR        = IQR(prices),
  range      = diff(range(prices)),
  CV_pct     = round(CV_pct, 1))
```

A CV in the $50$–$60\%$ range is substantial: the Granada AirBnB market is markedly heterogeneous.

### Comparing two markets with the CV

```{r ch01-cv-compare}
set.seed(99)
city_centre <- rnorm(50, mean = 100, sd = 15)
countryside <- rnorm(50, mean = 100, sd = 40)

cv <- function(x) sd(x) / mean(x) * 100
c(city_centre_CV = round(cv(city_centre), 1),
  countryside_CV = round(cv(countryside), 1))
```

Same mean, very different CV: the countryside market is far less homogeneous.

### Histogram with mean and median

```{r ch01-hist}
hist(prices, breaks = 20, col = "steelblue", border = "white",
     main = "AirBnB nightly prices in Granada (n = 80)",
     xlab = "Price (€)", las = 1)
abline(v = mean(prices),   col = "red",       lwd = 2, lty = 2)
abline(v = median(prices), col = "darkgreen", lwd = 2, lty = 2)
legend("topright", legend = c("Mean", "Median"),
       col = c("red", "darkgreen"), lwd = 2, lty = 2)
```

The red mean line sits to the right of the green median line — the classic visual signature of positive skewness.

### Skewness and excess kurtosis

```{r ch01-skew-kurt}
m3 <- mean((prices - mean(prices))^3)
m4 <- mean((prices - mean(prices))^4)
g1 <- m3 / S^3
g2 <- m4 / S^4 - 3

c(skewness_g1 = round(g1, 3),
  ex_kurtosis_g2 = round(g2, 3))
```

Positive $g_1$ confirms the right tail; positive $g_2$ confirms heavier-than-normal tails — extreme nightly prices are more likely here than a bell curve would predict.

### Lorenz curve and Gini index

```{r ch01-lorenz}
sorted    <- sort(prices)
pop_share <- (1:n) / n
inc_share <- cumsum(sorted) / sum(sorted)

plot(c(0, pop_share), c(0, inc_share), type = "l",
     col = "steelblue", lwd = 2,
     xlab = "Cumulative share of apartments",
     ylab = "Cumulative share of total revenue",
     main = "Lorenz curve — AirBnB Granada", las = 1)
abline(0, 1, col = "grey50", lty = 2)
legend("topleft", legend = c("Lorenz curve", "Perfect equality"),
       col = c("steelblue", "grey50"), lwd = 2, lty = c(1, 2))

# Gini via the trapezoidal rule.
B    <- sum((inc_share[-1] + inc_share[-n]) * diff(pop_share)) / 2
gini <- 1 - 2 * B
round(gini, 3)
```

A Gini around $0.30$ signals moderate concentration: the most expensive listings absorb a disproportionate share of total nightly revenue, but the market is far from one-firm dominance. For reference, the Spanish national income Gini reported by INE 2023 is $0.327$ — strikingly close to the Granada AirBnB price Gini in this small sample.

## Self-check {.unnumbered}

::: {.callout-tip collapse="true"}
## Q1. Absolute frequency

In a grouped frequency table, the absolute frequency $n_i$ of class $i$ is:

- A. The proportion of observations in class $i$.
- B. The number of observations falling inside class $i$.
- C. The cumulative number of observations up to and including class $i$.
- D. The midpoint of class $i$ multiplied by its width.

**Answer: B.** Absolute frequencies are raw counts; proportions are *relative* frequencies, cumulative counts are $N_i$, and midpoints are class marks.
:::

::: {.callout-tip collapse="true"}
## Q2. Relative frequencies sum to one

Given $n_i$ and total sample size $n$, the relative frequency is $f_i = n_i / n$. Which identity ALWAYS holds?

- A. $\sum_i f_i = 1$.
- B. $\sum_i f_i = n$.
- C. $f_i \leq n_i$ only if $n < 1$.
- D. $f_i$ is always larger than $n_i$.

**Answer: A.** Dividing every frequency by $n$ makes the relative frequencies a probability-mass-like vector that sums to one.
:::

::: {.callout-tip collapse="true"}
## Q3. Last cumulative frequency

The cumulative absolute frequency $N_i = \sum_{j \leq i} n_j$. The value of $N_k$ for the last class $k$ is:

- A. Equal to $1$.
- B. Equal to the largest absolute frequency.
- C. Equal to $n$, the total number of observations.
- D. Always equal to the number of classes.

**Answer: C.** All observations have been counted by the last class, so $N_k = n$.
:::

::: {.callout-tip collapse="true"}
## Q4. The mean as a minimiser

The arithmetic mean of $\{x_1, \ldots, x_n\}$ satisfies which optimisation property?

- A. It is the value $c$ that minimises $\sum_i |x_i - c|$.
- B. It is the value $c$ that minimises $\sum_i (x_i - c)^2$.
- C. It is always equal to the median.
- D. It is robust to extreme outliers.

**Answer: B.** Minimising squared deviations gives the mean; minimising absolute deviations gives the median.
:::

::: {.callout-tip collapse="true"}
## Q5. AM–GM inequality

For a strictly positive sample, the geometric mean $G = (\prod x_i)^{1/n}$ satisfies:

- A. $G \geq \bar{x}$ in any sample.
- B. $G \leq \bar{x}$, with equality only when all $x_i$ are equal.
- C. $G = \bar{x}$ if and only if the sample is symmetric.
- D. $G$ can be negative when some $x_i$ are large.

**Answer: B.** This is the AM–GM inequality. Equality holds only for constant samples.
:::

::: {.callout-tip collapse="true"}
## Q6. Effect of an outlier

Adding a single very large outlier (e.g. a €5,000 penthouse to the AirBnB sample) typically:

- A. Shifts the median substantially but barely moves the mean.
- B. Shifts the mean substantially but barely moves the median.
- C. Leaves both the mean and the median unchanged.
- D. Increases the median above the mean.

**Answer: B.** The mean responds to every observation; the median is a positional statistic and is essentially unmoved by a single extreme value.
:::

::: {.callout-tip collapse="true"}
## Q7. Linear transformation

If we transform every observation as $Y_i = a + b X_i$, then:

- A. $\bar{Y} = a + b\bar{X}$ and $S_Y = |b|\, S_X$.
- B. $\bar{Y} = b\bar{X}$ and $S_Y = b^2 S_X$.
- C. $\bar{Y} = a + \bar{X}$ and $S_Y = S_X + |b|$.
- D. Both the mean and the standard deviation are shifted by $a$.

**Answer: A.** The mean is fully linear (responds to both $a$ and $b$); the standard deviation only responds to $b$ in absolute value, so adding a constant does not change spread.
:::

::: {.callout-tip collapse="true"}
## Q8. Skewness sign

The Fisher skewness coefficient $g_1 = m_3 / S^3$ is positive when:

- A. The distribution has a left tail (a few small values pulling the mean below the median).
- B. The distribution has a right tail (a few large values pulling the mean above the median).
- C. The distribution is perfectly symmetric.
- D. The distribution has heavier tails than a normal.

**Answer: B.** Positive skewness = right tail = mean above median. Heavy tails are diagnosed by kurtosis ($g_2$), not skewness.
:::

::: {.callout-tip collapse="true"}
## Q9. Gini index interpretation

The Gini index $G \in [0, 1]$. Which interpretation is correct?

- A. $G = 1$ means perfect equality.
- B. $G = 0$ means perfect equality; values close to 1 mean almost all the total is held by one unit.
- C. $G$ can be negative if income is very dispersed.
- D. $G$ equals the Lorenz curve evaluated at $p = 0.5$.

**Answer: B.** The Gini is $0$ when everyone is equal (Lorenz curve = diagonal) and tends to $1$ when the entire total is concentrated in a single unit.
:::

## Exercises {.unnumbered}

### Exercise 1.1 ★ — Frequency table from raw data

The ages of 20 students in a class are
$$
18,\, 19,\, 19,\, 20,\, 20,\, 20,\, 20,\, 21,\, 21,\, 21,\, 21,\, 21,\, 22,\, 22,\, 22,\, 23,\, 23,\, 24,\, 25,\, 27.
$$

(a) Build a complete frequency table ($n_i$, $f_i$, $N_i$, $F_i$).
(b) What percentage of students are 21 or younger?
(c) Identify the mode and describe the shape qualitatively.

::: {.callout-tip collapse="true"}
## Solution

(a)

| $x_i$ | $n_i$ | $f_i$ | $N_i$ | $F_i$ |
|:---:|:---:|:---:|:---:|:---:|
| 18 | 1 | 0.05 | 1 | 0.05 |
| 19 | 2 | 0.10 | 3 | 0.15 |
| 20 | 4 | 0.20 | 7 | 0.35 |
| 21 | 5 | 0.25 | 12 | 0.60 |
| 22 | 3 | 0.15 | 15 | 0.75 |
| 23 | 2 | 0.10 | 17 | 0.85 |
| 24 | 1 | 0.05 | 18 | 0.90 |
| 25 | 1 | 0.05 | 19 | 0.95 |
| 27 | 1 | 0.05 | 20 | 1.00 |

(b) $F(21) = 0.60 = 60\%$.

(c) Mode $= 21$ (highest frequency, $n = 5$). The distribution is slightly right-skewed (the lonely $27$ stretches the right tail).
:::

### Exercise 1.2 ★ — Mean, variance, and CV

The weekly sales (in thousands of €) of a small business over 8 weeks are
$$
12,\, 15,\, 11,\, 14,\, 18,\, 13,\, 16,\, 17.
$$
Compute (a) the mean, (b) the variance, (c) the standard deviation, (d) the coefficient of variation.

::: {.callout-tip collapse="true"}
## Solution

(a) $\sum x_i = 116$, $\bar{x} = 116/8 = 14.5$ thousand €.

(b) $\sum x_i^2 = 1724$, $a_2 = 1724/8 = 215.5$. Hence $S^2 = 215.5 - 14.5^2 = 215.5 - 210.25 = 5.25$.

(c) $S = \sqrt{5.25} \approx 2.291$ thousand €.

(d) $CV = 2.291 / 14.5 \approx 0.158$. Low relative dispersion — the mean is quite representative.
:::

### Exercise 1.3 ★ — Linear transformation: Celsius → Fahrenheit

Five daily temperatures (in $^\circ$C) are $15, 18, 22, 20, 25$. Let $F = 32 + 1.8 C$.

(a) Compute $\bar{x}$ and $S$ in Celsius.
(b) Compute $\bar{y}$ and $S_Y$ in Fahrenheit using the linear-transformation rules.
(c) Verify directly from the Fahrenheit observations.

::: {.callout-tip collapse="true"}
## Solution

(a) $\bar{x} = 100/5 = 20$. $\sum x_i^2 = 2058$, so $S^2 = 2058/5 - 400 = 11.6$ and $S \approx 3.406\,^\circ$C.

(b) $\bar{y} = 32 + 1.8 \times 20 = 68\,^\circ$F. $S_Y = |1.8|\, S_X = 1.8 \times 3.406 \approx 6.131\,^\circ$F.

(c) Fahrenheit data: $59,\, 64.4,\, 71.6,\, 68,\, 77$. Direct calculation gives $\bar{y} = 340/5 = 68$ and $S_Y \approx 6.131$ — matches part (b).
:::

### Exercise 1.4 ★★ — Grouped data: full descriptive analysis

The monthly electricity bills (€) for 60 households are summarised in the table below.

| Bill (€) | $n_i$ |
|:---:|:---:|
| $[20, 40)$ | 8 |
| $[40, 60)$ | 15 |
| $[60, 80)$ | 20 |
| $[80, 100)$ | 12 |
| $[100, 120)$ | 5 |

Compute (a) the mean, (b) the median, (c) the variance and standard deviation, (d) $Q_1$ and $Q_3$, (e) the coefficient of variation. Comment briefly on the symmetry of the distribution.

### Exercise 1.5 ★★ — Comparing two chains with the CV

Two supermarket chains report weekly sales (in thousands of €) over 8 weeks:

| Week | Chain A | Chain B |
|:---:|:---:|:---:|
| 1 | 42 | 120 |
| 2 | 38 | 135 |
| 3 | 45 | 110 |
| 4 | 40 | 128 |
| 5 | 50 | 145 |
| 6 | 35 | 115 |
| 7 | 43 | 140 |
| 8 | 47 | 130 |

(a) Compute the mean, variance, and standard deviation of each chain.
(b) Compute the CV of each chain.
(c) Which chain shows greater *relative* dispersion? Which mean is more representative?

### Exercise 1.6 ★★ — Skewness from grouped data

The ages of 50 participants in a training programme are grouped as follows.

| Age | $n_i$ |
|:---:|:---:|
| $[20, 25)$ | 4 |
| $[25, 30)$ | 10 |
| $[30, 35)$ | 18 |
| $[35, 40)$ | 12 |
| $[40, 45]$ | 6 |

(a) Compute the mean, median, and mode.
(b) Compute Fisher's skewness $g_1 = m_3 / S^3$.
(c) Interpret the result.

### Exercise 1.7 ★★ — Quartiles, IQR, and box-plot description

Sixty bank branches report their daily number of transactions in the following table.

| Transactions | $n_i$ |
|:---:|:---:|
| $[20, 40)$ | 8 |
| $[40, 60)$ | 15 |
| $[60, 80)$ | 20 |
| $[80, 100)$ | 12 |
| $[100, 120]$ | 5 |

(a) Compute $Q_1$, $Q_2$, and $Q_3$.
(b) Compute the IQR and the inner fences $Q_1 - 1.5\,IQR$ and $Q_3 + 1.5\,IQR$. Are there potential outliers?
(c) Describe the appearance of a box-plot of this distribution.

### Exercise 1.8 ★★★ — Lorenz curve and Gini index

Two hundred households are grouped by annual income (thousands of €).

| Income | $n_i$ |
|:---:|:---:|
| $[8, 16)$ | 30 |
| $[16, 24)$ | 50 |
| $[24, 32)$ | 60 |
| $[32, 40)$ | 40 |
| $[40, 48]$ | 20 |

(a) Compute $(p_i, q_i)$ for each interval (class marks as representative values).
(b) Sketch the Lorenz curve.
(c) Compute the Gini index using the formula $G = 1 - \sum (p_i - p_{i-1})(q_i + q_{i-1})$.
(d) Interpret the result.

### Exercise 1.9 ★★★ — Comprehensive analysis from raw data

Twenty-five supermarkets in Málaga charge the following prices (€) for a standard basket of groceries:
$$
32, 28, 35, 41, 30, 37, 29, 33, 45, 38, 31, 34, 40,
$$
$$
27, 36, 33, 42, 30, 35, 39, 28, 34, 37, 43, 31.
$$

(a) Build a frequency table using the intervals $[27,31)$, $[31,35)$, $[35,39)$, $[39,43)$, $[43,47)$.
(b) Compute the mean, median, and mode.
(c) Compute the variance, standard deviation, and coefficient of variation.
(d) Compute $Q_1$, $Q_3$, and the IQR.
(e) Compute the Pearson skewness coefficient $A_p = 3(\bar{x} - Me)/S$ and interpret.
(f) If a 7% VAT is added (new price $= 1.07 \times$ old price), what are the new mean and standard deviation?
