---
title: "Introduction to Probability"
---


## Learning outcomes {.unnumbered}

By the end of this chapter the reader should be able to:

- Define a random experiment, sample space, and event, and represent events as subsets of $\Omega$.
- Combine events using the standard set operations (union, intersection, complement, difference) and apply De Morgan's laws.
- State Kolmogorov's three axioms and derive the complement rule, monotonicity, and the inclusion-exclusion formula.
- Compute classical (Laplace) probabilities by counting equally likely outcomes.
- Compute conditional probabilities $P(A \mid B)$ and apply the multiplication rule, including in tree-diagram problems.
- Apply the law of total probability and Bayes' theorem to update beliefs from prior to posterior given evidence.
- Distinguish independence from mutual exclusivity, and recognise common pitfalls (e.g. the base-rate fallacy).
- Simulate dice, coins, the birthday paradox, a medical test, and the Monty Hall problem in R.

## Motivating empirical question {.unnumbered}

> *A rapid test for a disease that affects 1% of the population is 99% sensitive and 95% specific. You test positive. What is the probability that you are actually sick?*

The intuitive answer "about 99%" is wrong by an order of magnitude --- the correct answer is about 17%. Probability theory, and in particular Bayes' theorem, is the machinery that turns prior beliefs (the prevalence) and conditional evidence (the test result) into a posterior probability. The same logic governs credit-default models, fraud detection, recession indicators, and insurance pricing. This chapter builds that machinery from the ground up, starting with sample spaces and ending with the false-positive paradox and the Monty Hall problem.

## 3.1 Random experiments and sample spaces

`\lettrine[lines=3]{I}{n}`{=latex} the first two chapters we *described* data that had already been collected. We now shift to **modelling uncertainty**: how should we reason about outcomes that have not yet occurred? The mathematical framework is **probability theory**.

::: {.callout-note}
## Definition: random experiment

A **random experiment** (or *aleatory experiment*) is any process satisfying three conditions:

(i) All possible outcomes can be specified *before* the experiment is performed.
(ii) The actual outcome *cannot* be predicted with certainty in advance.
(iii) The experiment can, in principle, be repeated under the same conditions.
:::

Familiar examples include rolling a die, tossing a coin, tomorrow's closing price of the IBEX-35, and the number of customers entering a shop in one hour. A deterministic process such as evaluating $2+3$ is *not* a random experiment because the outcome is fixed.

::: {.callout-note}
## Definition: sample space

The **sample space** $\Omega$ is the set of all possible outcomes of a random experiment. Each $\omega \in \Omega$ is an **elementary outcome** (or *sample point*).
:::

A sample space is **discrete** if it is finite or countably infinite, and **continuous** if it is uncountable. In this chapter we work almost exclusively with discrete (and usually finite) sample spaces; continuous sample spaces appear systematically from Chapter 4 onwards.

The following table records a few canonical examples.

| Experiment | Sample space $\Omega$ | $\lvert\Omega\rvert$ |
|---|---|---|
| Toss a coin | $\{H, T\}$ | 2 |
| Roll a six-sided die | $\{1, 2, 3, 4, 5, 6\}$ | 6 |
| Toss a coin twice | $\{HH, HT, TH, TT\}$ | 4 |
| Roll two dice | $\{(i,j) : i,j \in \{1,\ldots,6\}\}$ | 36 |
| Calls to a call centre in a day | $\{0, 1, 2, \ldots\}$ | $\infty$ (countable) |
| Waiting time at a bus stop (min) | $[0, \infty)$ | $\infty$ (uncountable) |

## 3.2 Events and operations on events

::: {.callout-note}
## Definition: event

An **event** is any subset $A \subseteq \Omega$. We say $A$ **occurs** if the realised outcome $\omega$ belongs to $A$.
:::

Three special kinds of events recur throughout the chapter:

- An **elementary event** is a singleton $\{\omega\}$.
- The **sure event** is $\Omega$ itself; it always occurs.
- The **impossible event** is the empty set $\emptyset$; it never occurs.

Since events are sets, we combine them with the standard set operations.

::: {.callout-note}
## Definition: set operations on events

For $A, B \subseteq \Omega$:

- **Union:** $A \cup B = \{\omega : \omega \in A \text{ or } \omega \in B\}$ ("$A$ or $B$ occurs").
- **Intersection:** $A \cap B = \{\omega : \omega \in A \text{ and } \omega \in B\}$ ("both occur").
- **Complement:** $\bar{A} = \Omega \setminus A$ ("$A$ does not occur").
- **Difference:** $A \setminus B = A \cap \bar{B}$ ("$A$ but not $B$").
- Two events are **mutually exclusive** (or *disjoint*) if $A \cap B = \emptyset$.
:::

::: {.callout-note}
## Theorem: De Morgan's laws

For any events $A$ and $B$:
:::

$$
\overline{A \cup B} = \bar{A} \cap \bar{B}, \qquad \overline{A \cap B} = \bar{A} \cup \bar{B}.
$$

The laws generalise to any finite (or countably infinite) collection of events. In words: the complement of a union is the intersection of complements, and the complement of an intersection is the union of complements.

::: {.callout-note}
## Example: set operations on a die

Roll a die and let $A = \{2, 4, 6\}$ (even) and $B = \{1, 2, 3\}$ (at most 3). Then $A \cup B = \{1,2,3,4,6\}$, $A \cap B = \{2\}$, $\bar{A} = \{1,3,5\}$, and $A \setminus B = \{4, 6\}$. The events are *not* mutually exclusive because $A \cap B \ne \emptyset$.
:::

## 3.3 Axioms of probability

The modern axiomatic foundation was set down by Kolmogorov in 1933. The entire theory rests on three rules.

::: {.callout-note}
## Definition: Kolmogorov's axioms

A **probability function** $P$ on $\Omega$ assigns a real number $P(A)$ to each event $A \subseteq \Omega$ and satisfies:

(A1) **Non-negativity:** $P(A) \geq 0$ for every event $A$.
(A2) **Normalisation:** $P(\Omega) = 1$.
(A3) **Additivity:** if $A \cap B = \emptyset$, then $P(A \cup B) = P(A) + P(B)$. More generally, if $A_1, A_2, \ldots$ are pairwise disjoint, $P\!\left(\bigcup_i A_i\right) = \sum_i P(A_i)$.
:::

From these three axioms one can derive every other property of probability. The most useful are collected below.

::: {.callout-note}
## Theorem: derived properties

For any events $A, B \subseteq \Omega$:

(i) **Complement rule:** $P(\bar{A}) = 1 - P(A)$.
(ii) **Impossible event:** $P(\emptyset) = 0$.
(iii) **Bound:** $0 \leq P(A) \leq 1$.
(iv) **Monotonicity:** if $A \subseteq B$ then $P(A) \leq P(B)$.
(v) **Inclusion-exclusion (two events):**
:::

$$
P(A \cup B) = P(A) + P(B) - P(A \cap B).
$$

The three-event version is

$$
\begin{aligned}
P(A \cup B \cup C) &= P(A) + P(B) + P(C) \\
&\quad - P(A \cap B) - P(A \cap C) - P(B \cap C) + P(A \cap B \cap C).
\end{aligned}
$$

**Proof sketch of the complement rule.** $A$ and $\bar{A}$ are disjoint with $A \cup \bar{A} = \Omega$, so by (A3) and (A2), $P(A) + P(\bar{A}) = 1$.

::: {.callout-note}
## Example: inclusion-exclusion in a language survey

Among 200 economics students, 120 study English, 80 study French, and 30 study both. The probability that a randomly chosen student studies *at least one* of the two languages is
:::

$$
P(E \cup F) = \frac{120}{200} + \frac{80}{200} - \frac{30}{200} = \frac{170}{200} = 0.85.
$$

So 85% of the students study at least one language; the remaining 15% study neither.

## 3.4 Classical (Laplace) probability

::: {.callout-note}
## Definition: classical (Laplace) probability

If $\Omega$ is finite and all elementary outcomes are equally likely, then for any event $A$:
:::

$$
P(A) = \frac{|A|}{|\Omega|} = \frac{\text{number of favourable outcomes}}{\text{total number of outcomes}}.
$$

::: {.callout-warning}
## Common mistake: assuming equally likely without justification

The Laplace rule applies *only* when symmetry justifies equal likelihood (fair coins, fair dice, a well-shuffled deck, a fair lottery). If outcomes are not symmetric --- e.g. the daily return of a stock, or the lifetime of a light bulb --- the formula does not apply and probabilities must be estimated from data or modelled separately.
:::

::: {.callout-note}
## Example: rolling two fair dice

The sample space has $\lvert\Omega\rvert = 36$ equally likely pairs. Let $A$ = "sum is 7" and $B$ = "sum is 12". Then $A = \{(1,6),(2,5),(3,4),(4,3),(5,2),(6,1)\}$ gives $P(A) = 6/36 = 1/6 \approx 0.167$, and $B = \{(6,6)\}$ gives $P(B) = 1/36 \approx 0.028$. The event $C$ = "at least one die shows a 6" is easiest computed by complement: there are $5 \times 5 = 25$ outcomes with no six, so $P(C) = 1 - 25/36 = 11/36 \approx 0.306$.
:::

Counting arguments such as these rely on the basic combinatorial tools (permutations, combinations, binomial coefficients). A self-contained refresher lives in [Appendix A](appendix-a-prerequisites.qmd); here we use the formulas as needed.

## 3.5 Conditional probability

Partial information about the outcome often changes our assessment of probabilities. This leads to **conditional probability**.

::: {.callout-note}
## Definition: conditional probability

Let $A$ and $B$ be events with $P(B) > 0$. The **conditional probability of $A$ given $B$** is
:::

$$
P(A \mid B) = \frac{P(A \cap B)}{P(B)}.
$$

The intuition is that knowing $B$ has occurred shrinks the sample space from $\Omega$ to $B$; we then ask which fraction of the new universe also belongs to $A$.

::: {.callout-note}
## Example: conditional probability on a die

Roll a fair die with $A = \{2,4,6\}$ (even) and $B = \{4,5,6\}$ (greater than 3). Then $A \cap B = \{4, 6\}$ and
:::

$$
P(A \mid B) = \frac{P(A \cap B)}{P(B)} = \frac{2/6}{3/6} = \frac{2}{3}.
$$

Knowing the die showed a number greater than 3 raises the probability of an even result from $1/2$ to $2/3$.

### 3.5.1 The multiplication rule

Rearranging the definition gives the **multiplication rule**:

$$
P(A \cap B) = P(A \mid B)\, P(B) = P(B \mid A)\, P(A).
$$

The generalisation to $n$ events is the chain rule

$$
P(A_1 \cap \cdots \cap A_n) = P(A_1)\, P(A_2 \mid A_1)\, P(A_3 \mid A_1 \cap A_2) \cdots P(A_n \mid A_1 \cap \cdots \cap A_{n-1}).
$$

::: {.callout-note}
## Example: drawing balls without replacement

An urn contains 5 red and 3 blue balls. Two balls are drawn without replacement. Let $R_1, R_2$ be "first red" and "second red." Then
:::

$$
P(R_1 \cap R_2) = P(R_1)\, P(R_2 \mid R_1) = \frac{5}{8} \cdot \frac{4}{7} = \frac{20}{56} = \frac{5}{14} \approx 0.357.
$$

The four terminal outcomes RR, RB, BR, BB have probabilities $20/56,\,15/56,\,15/56,\,6/56$, which sum to $1$ as required.

### 3.5.2 Independence

::: {.callout-note}
## Definition: independent events

Events $A$ and $B$ are **independent** if
:::

$$
P(A \cap B) = P(A) \, P(B).
$$

Equivalently, when $P(B) > 0$, independence means $P(A \mid B) = P(A)$: knowing $B$ does not change the probability of $A$.

::: {.callout-warning}
## Common mistake: independence is *not* mutual exclusivity

Two events are **mutually exclusive** if $A \cap B = \emptyset$; they cannot occur together. Two events are **independent** if $P(A \cap B) = P(A)\, P(B)$; one does not affect the other. If $A$ and $B$ are mutually exclusive with $P(A) > 0$ and $P(B) > 0$, then $P(A \cap B) = 0 \ne P(A)\,P(B)$, so they are *not* independent --- in fact, mutually exclusive events are *maximally dependent*: observing one guarantees the other did not occur.
:::

::: {.callout-note}
## Example: two coin tosses

Toss a fair coin twice and let $A$ = "first toss is Heads", $B$ = "second toss is Heads". Then $P(A) = P(B) = 1/2$ and $P(A \cap B) = P(\{HH\}) = 1/4 = (1/2)(1/2)$. So $A$ and $B$ are independent, matching intuition.
:::

## 3.6 The birthday paradox

A famous illustration of how badly intuition can mislead us is the **birthday problem**.

> *In a room of $n$ people chosen at random, what is the probability that at least two share a birthday?*

Assume birthdays are independent and uniform on 365 days (ignoring leap years). It is easier to compute the complement $\bar{A}$ = "all birthdays distinct":

$$
P(\bar{A}) = \frac{365}{365} \cdot \frac{364}{365} \cdots \frac{365 - n + 1}{365} = \prod_{k=0}^{n-1} \frac{365 - k}{365},
$$

so

$$
P(A) = 1 - \prod_{k=0}^{n-1} \frac{365 - k}{365}.
$$

| $n$ | $P(\text{match})$ | $n$ | $P(\text{match})$ |
|---|---|---|---|
| 5 | 0.027 | 30 | 0.706 |
| 10 | 0.117 | 40 | 0.891 |
| **23** | **0.507** | 50 | 0.970 |
| 25 | 0.569 | 60 | 0.994 |

With only $n = 23$ people the probability exceeds 50%. The trap is that we mentally compare 365 days to the number of people, when the relevant count is the number of *pairs*: with 23 people there are $\binom{23}{2} = 253$ pairs, each with a small but non-trivial chance of matching. Pairwise counts grow quadratically, which is why the curve climbs so fast.

## 3.7 Law of total probability

Many problems are easier to attack by first splitting $\Omega$ into a partition and then summing the conditional probabilities.

::: {.callout-note}
## Definition: partition

Events $A_1, \ldots, A_n$ form a **partition** of $\Omega$ if (i) they are pairwise disjoint and (ii) their union is $\Omega$. Every outcome belongs to exactly one $A_i$.
:::

::: {.callout-note}
## Theorem: law of total probability

If $\{A_1, \ldots, A_n\}$ is a partition of $\Omega$ with $P(A_i) > 0$, then for any event $B$,
:::

$$
P(B) = \sum_{i=1}^{n} P(B \mid A_i)\, P(A_i).
$$

The idea: $B$ decomposes into the disjoint pieces $B \cap A_1, \ldots, B \cap A_n$, and additivity gives the sum.

::: {.callout-note}
## Example: supplier defect rates

A factory receives components from three suppliers: $A_1$ supplies 50% with a 2% defect rate, $A_2$ supplies 30% with a 3% defect rate, and $A_3$ supplies 20% with a 5% defect rate. Let $D$ = "defective." The law of total probability gives
:::

$$
P(D) = (0.02)(0.50) + (0.03)(0.30) + (0.05)(0.20) = 0.010 + 0.009 + 0.010 = 0.029.
$$

So 2.9% of all components are defective. The same calculation works for insurance claim rates, recession indicators, and any other "different sub-populations have different rates" problem.

## 3.8 Bayes' theorem

Bayes' theorem *inverts* a conditional probability: given $P(B \mid A_i)$, it returns $P(A_i \mid B)$. It is the engine of every diagnostic test, every spam filter, and every credit-scoring algorithm.

::: {.callout-note}
## Theorem: Bayes' theorem

If $\{A_1, \ldots, A_n\}$ partition $\Omega$ with $P(A_i) > 0$, and $P(B) > 0$, then
:::

$$
P(A_i \mid B) = \frac{P(B \mid A_i)\, P(A_i)}{\displaystyle\sum_{j=1}^{n} P(B \mid A_j)\, P(A_j)} = \frac{P(B \mid A_i)\, P(A_i)}{P(B)}.
$$

The terminology is standard:

- $P(A_i)$ is the **prior** — belief about $A_i$ before observing the evidence.
- $P(B \mid A_i)$ is the **likelihood** — the probability of the evidence under each hypothesis.
- $P(A_i \mid B)$ is the **posterior** — the updated belief after observing $B$.

In words: $\text{posterior} = \dfrac{\text{likelihood} \times \text{prior}}{\text{evidence}}.$

### 3.8.1 The false-positive paradox

The Bayesian framework reveals an effect that is genuinely surprising the first time one sees it: a highly accurate test for a *rare* condition can still produce mostly false positives.

::: {.callout-note}
## Definition: test characteristics

For a binary test of a disease:

- **Sensitivity** (true positive rate): $P(+ \mid \text{sick})$.
- **Specificity** (true negative rate): $P(- \mid \text{healthy})$.
- **Prevalence**: $P(\text{sick})$.
- **Positive predictive value (PPV)**: $P(\text{sick} \mid +)$.
:::

::: {.callout-note}
## Example: a 99% accurate test for a 1% disease

A disease has prevalence 1%. A test has sensitivity 99% and specificity 99% (so $P(+ \mid \text{healthy}) = 0.01$). If a randomly selected person tests positive, what is the probability they are sick?
:::

By the law of total probability,

$$
P(+) = (0.99)(0.01) + (0.01)(0.99) = 0.0099 + 0.0099 = 0.0198,
$$

and by Bayes' theorem,

$$
P(\text{sick} \mid +) = \frac{(0.99)(0.01)}{0.0198} = 0.50.
$$

Even with a 99% accurate test, a positive result corresponds to only a **50% chance** of being sick. The frequency-table version makes this concrete: out of 10,000 people, 100 are sick and 9,900 healthy. The test catches 99 of the 100 sick (true positives) but also flags $0.01 \times 9{,}900 = 99$ healthy people (false positives). True and false positives are equal in number, so $P(\text{sick} \mid +) = 99/(99+99) = 1/2$.

| | Test $+$ | Test $-$ | Total |
|---|---|---|---|
| Sick (1%) | 99 | 1 | 100 |
| Healthy (99%) | 99 | 9,801 | 9,900 |
| **Total** | **198** | **9,802** | **10,000** |

The PPV depends critically on the prevalence:

| Prevalence | PPV |
|---|---|
| 0.1% | 9.0% |
| 0.5% | 33.2% |
| 1% | 50.0% |
| 2% | 66.9% |
| 5% | 83.9% |
| 10% | 91.7% |
| 20% | 96.1% |
| 50% | 99.0% |

The same logic governs fraud detection (a 99% accurate fraud filter swamps the analyst with false alarms if the true fraud rate is low), spam classification, and credit-risk scoring. **The base rate is the dominant input** — without it the test's accuracy alone is meaningless.

### 3.8.2 Bayes in economics: market testing

::: {.callout-note}
## Example: market testing a new product

A company knows that 40% of similar products succeed ($S$) and 60% fail ($F$). A pre-launch market test gives a positive signal 80% of the time when the product will succeed and 30% of the time when it will fail. If the test is positive, what is the posterior probability of success?
:::

First the evidence:

$$
P(+) = (0.80)(0.40) + (0.30)(0.60) = 0.32 + 0.18 = 0.50.
$$

Then Bayes:

$$
P(S \mid +) = \frac{(0.80)(0.40)}{0.50} = 0.64.
$$

The positive test result raises the success probability from a prior of 40% to a posterior of 64% — useful, but far from a guarantee.

## 3.9 The Monty Hall problem

The **Monty Hall problem**, named after the host of *Let's Make a Deal*, is the most famous demonstration that conditioning on a *non-random* event can violently shift probabilities.

> *Three doors hide one car and two goats. You pick a door. The host, who knows where the car is, opens a different door to reveal a goat and offers you the chance to switch. Should you switch?*

Let $C_i$ = "car is behind door $i$" with $P(C_1) = P(C_2) = P(C_3) = 1/3$. Suppose you pick door 1 and the host opens door 3, revealing a goat (event $H_3$). The host's behaviour is constrained: he never opens your door and never opens the car door, so

$$
P(H_3 \mid C_1) = \tfrac{1}{2}, \qquad P(H_3 \mid C_2) = 1, \qquad P(H_3 \mid C_3) = 0.
$$

The total probability of $H_3$ is

$$
P(H_3) = \tfrac{1}{3} \cdot \tfrac{1}{2} + \tfrac{1}{3} \cdot 1 + \tfrac{1}{3} \cdot 0 = \tfrac{1}{2}.
$$

Bayes' theorem gives

$$
P(C_1 \mid H_3) = \frac{(1/3)(1/2)}{1/2} = \frac{1}{3}, \qquad P(C_2 \mid H_3) = \frac{(1/3)(1)}{1/2} = \frac{2}{3}.
$$

So *staying* wins with probability $1/3$ and *switching* wins with probability $2/3$. **You should always switch.** The host's action is informative precisely because it is constrained — your initial pick had probability $1/3$ of being right and nothing the host does later changes that, so the remaining $2/3$ collapses onto the other unopened door.

The naive "two doors left, so 50/50" intuition treats the host's choice as random. It isn't. This is a textbook example of how conditioning on a non-random observation reshapes posterior beliefs.

## 3.10 R Lab — Simulating probability with the law of large numbers

Probability theory can feel abstract, but simulation makes it tangible. By running thousands of random experiments on a computer we can watch empirical frequencies converge to theoretical probabilities — the **law of large numbers** in action. Throughout this lab we set `set.seed(2026)` so results are reproducible across the whole book.

```{r ch03-setup}
#| message: false
#| warning: false
set.seed(2026)
n_sim <- 10000
```

### Dice and coins

```{r ch03-dice}
rolls <- sample(1:6, n_sim, replace = TRUE)
freq  <- table(rolls) / n_sim
round(freq, 4)
```

Each face appears close to $1/6 \approx 0.1667$. With more rolls the deviations shrink further.

```{r ch03-coin}
flips <- sample(c("Heads", "Tails"), n_sim, replace = TRUE)
round(table(flips) / n_sim, 4)
```

The **law of large numbers** says that the running average of i.i.d.\ draws converges to the expected value. For a fair die, $\mathbb{E}[X] = 3.5$.

```{r ch03-lln-plot}
running <- cumsum(rolls) / seq_len(n_sim)
plot(seq_len(n_sim), running, type = "l", col = "steelblue",
     xlab = "Number of rolls", ylab = "Running average",
     main = "Law of large numbers - die-roll average",
     las = 1, ylim = c(2.5, 4.5))
abline(h = 3.5, col = "red", lty = 2, lwd = 2)
legend("topright", legend = c("Running average", "E[X] = 3.5"),
       col = c("steelblue", "red"), lty = c(1, 2), lwd = 2)
```

The running average wobbles wildly at first but settles tightly around 3.5 as $n$ grows.

### Birthday paradox

```{r ch03-birthday-fun}
birthday_sim <- function(n_people, n_sim = 10000) {
  matches <- replicate(n_sim, {
    bdays <- sample(1:365, n_people, replace = TRUE)
    any(duplicated(bdays))
  })
  mean(matches)
}

p23 <- birthday_sim(23)
c(simulated = round(p23, 4), theoretical = 0.5073)
```

```{r ch03-birthday-curve}
group_sizes <- 2:60
probs <- sapply(group_sizes, birthday_sim)
plot(group_sizes, probs, type = "l", col = "steelblue", lwd = 2,
     xlab = "Number of people", ylab = "P(at least one shared birthday)",
     main = "Birthday paradox", las = 1)
abline(h = 0.5, col = "red", lty = 2)
abline(v = 23, col = "grey50", lty = 3)
```

By 50 people a match is almost certain. The number of pairs grows as $\binom{n}{2}$ — quadratically, not linearly.

### Conditional probability on cards

```{r ch03-cards}
suits <- rep(c("Hearts", "Diamonds", "Clubs", "Spades"), each = 13)
colors <- ifelse(suits %in% c("Hearts", "Diamonds"), "Red", "Black")
deck <- data.frame(Suit = suits, Color = colors, stringsAsFactors = FALSE)

draws <- deck[sample(52, n_sim, replace = TRUE), ]
red_draws <- draws[draws$Color == "Red", ]
c(sim = round(mean(red_draws$Suit == "Hearts"), 4),
  theory = round(13/26, 4))
```

Half the red cards are hearts, so $P(\text{Heart} \mid \text{Red}) = 1/2$, confirming $P(A \mid B) = P(A \cap B)/P(B)$.

### Bayes' theorem: the medical test

```{r ch03-bayes-sim}
n_pop       <- 1e6
prevalence  <- 0.01
sensitivity <- 0.99
specificity <- 0.95

disease     <- rbinom(n_pop, 1, prevalence)
test_result <- ifelse(disease == 1,
                      rbinom(n_pop, 1, sensitivity),
                      rbinom(n_pop, 1, 1 - specificity))

conf <- table(Disease = factor(disease,     labels = c("No",  "Yes")),
              Test    = factor(test_result, labels = c("Neg", "Pos")))
conf
```

```{r ch03-bayes-ppv}
tp <- conf["Yes", "Pos"]; fp <- conf["No", "Pos"]
ppv_sim    <- tp / (tp + fp)
ppv_theory <- (sensitivity * prevalence) /
              (sensitivity * prevalence + (1 - specificity) * (1 - prevalence))
c(simulated = round(ppv_sim, 4), bayes = round(ppv_theory, 4))
```

Despite 99% sensitivity, a positive result corresponds to only about a 17% chance of disease — the base-rate fallacy in action.

### Monty Hall

```{r ch03-monty}
simulate_monty <- function(n, strategy) {
  wins <- 0
  for (i in 1:n) {
    car  <- sample(1:3, 1)
    pick <- sample(1:3, 1)
    available  <- setdiff(1:3, c(pick, car))
    host_opens <- available[sample(length(available), 1)]
    final <- if (strategy == "switch") setdiff(1:3, c(pick, host_opens)) else pick
    if (final == car) wins <- wins + 1
  }
  wins / n
}

p_stay   <- simulate_monty(n_sim, "stay")
p_switch <- simulate_monty(n_sim, "switch")
c(stay = round(p_stay, 4), switch = round(p_switch, 4))
```

```{r ch03-monty-bar}
barplot(c(Stay = p_stay, Switch = p_switch),
        col = c("coral", "steelblue"), ylim = c(0, 0.8),
        main = "Monty Hall: stay vs switch", ylab = "Win probability", las = 1)
abline(h = c(1/3, 2/3), lty = 2, col = "grey50")
```

Switching wins roughly $2/3$ of the time, matching the Bayesian analysis above.

## Self-check {.unnumbered}

Seven short multiple-choice questions. Try each one before opening the answer.

::: {.callout-tip collapse="true"}
## Q1. The Kolmogorov axioms

The Kolmogorov axioms require, for any event $A$ in the sample space $\Omega$, that:

- A. $P(A) \in [-1, 1]$ and $P(\Omega) = 0$.
- B. $P(A) \geq 0$, $P(\Omega) = 1$, and $P(A \cup B) = P(A) + P(B)$ when $A$ and $B$ are mutually exclusive.
- C. $P(A) > 0$ for every event.
- D. $P(A \cup B) = P(A) \cdot P(B)$ for any pair of events.

**Answer: B.** These are the three axioms — non-negativity, normalisation, and finite additivity for disjoint events.
:::

::: {.callout-tip collapse="true"}
## Q2. Inclusion-exclusion

For two events $A$ and $B$, the formula $P(A \cup B) = P(A) + P(B) - P(A \cap B)$ subtracts $P(A \cap B)$ because:

- A. Otherwise the intersection $A \cap B$ would be counted twice.
- B. Probabilities must always sum to 1.
- C. It corrects for non-negativity violations.
- D. It is the definition of independence.

**Answer: A.** Adding $P(A)$ and $P(B)$ double-counts the overlap; subtracting $P(A \cap B)$ removes the duplication.
:::

::: {.callout-tip collapse="true"}
## Q3. Independence

Two events $A$ and $B$ are independent if and only if:

- A. $P(A \cap B) = P(A) \cdot P(B)$.
- B. $P(A \cup B) = P(A) + P(B)$.
- C. $A$ and $B$ are mutually exclusive.
- D. $P(A \mid B) = 0$.

**Answer: A.** This is the definition. Option C describes mutual exclusivity, which is in fact *incompatible* with independence when both events have positive probability.
:::

::: {.callout-tip collapse="true"}
## Q4. Conditional probability

The conditional probability of $A$ given $B$ (when $P(B) > 0$) is defined as:

- A. $P(A \mid B) = P(A) / P(B)$.
- B. $P(A \mid B) = P(A \cap B) / P(B)$.
- C. $P(A \mid B) = P(A) \cdot P(B)$.
- D. $P(A \mid B) = P(B \mid A)$ for any pair of events.

**Answer: B.** This is the textbook definition. Conditioning on $B$ rescales by the new sample space $B$.
:::

::: {.callout-tip collapse="true"}
## Q5. Total probability and Bayes

Bayes' theorem expands the denominator $P(B)$ via the law of total probability as:

- A. $P(B) = P(B \mid A) P(A) + P(B \mid A^c) P(A^c)$.
- B. $P(B) = P(B \mid A) + P(B \mid A^c)$.
- C. $P(B) = P(A) + P(A^c)$.
- D. $P(B) = 1 - P(A)$.

**Answer: A.** The partition is $\{A, A^c\}$; total probability adds the weighted conditional probabilities.
:::

::: {.callout-tip collapse="true"}
## Q6. The base-rate fallacy

In the medical test example (sensitivity 99%, specificity 95%, prevalence 1%) the PPV is $P(D \mid +) \approx 0.17$. The intuition is:

- A. With only 1% of the population sick, the 5% false-positive rate among the 99% healthy generates more false positives than true positives — the **base rate** dominates.
- B. 99% sensitivity guarantees a 99% probability of disease given a positive test.
- C. Specificity is irrelevant once prevalence is known.
- D. The PPV equals sensitivity minus the false-positive rate.

**Answer: A.** The product (false-positive rate)$\times$(prevalence of healthy) is the dominant term when the disease is rare.
:::

::: {.callout-tip collapse="true"}
## Q7. Monty Hall

In the Monty Hall problem (three doors, host opens a goat door after your first pick), switching wins with probability:

- A. $1/2$, because two doors remain.
- B. $1/3$, the same as staying.
- C. $2/3$, because the host's constrained action concentrates the remaining probability on the unopened door.
- D. $1$, switching always wins.

**Answer: C.** The host's behaviour is informative because he never opens your door and never opens the car door. The $2/3$ posterior collapses onto the door he leaves closed.
:::

## Exercises {.unnumbered}

### Exercise 3.1 ★ — Set operations on a die

A fair die is rolled once. Define $A = \{1,2,3\}$, $B = \{3,4,5,6\}$, $C = \{2,4,6\}$. Compute:

(a) $P(A \cup B)$.
(b) $P(A \cap C)$.
(c) $P(\bar{B} \cup C)$.
(d) $P(A \cap B \cap C)$.
(e) Are $A$ and $C$ independent?

::: {.callout-tip collapse="true"}
## Solution

(a) $A \cup B = \Omega$ so $P(A \cup B) = 1$.
(b) $A \cap C = \{2\}$ so $P(A \cap C) = 1/6$.
(c) $\bar{B} = \{1,2\}$, $\bar{B} \cup C = \{1,2,4,6\}$, so $P(\bar{B} \cup C) = 4/6 = 2/3$.
(d) $A \cap B = \{3\}$ and $\{3\} \cap C = \emptyset$, so $P(A \cap B \cap C) = 0$.
(e) $P(A)P(C) = (1/2)(1/2) = 1/4 \ne 1/6 = P(A \cap C)$, so $A$ and $C$ are **not independent**.
:::

### Exercise 3.2 ★ — Laplace probability with a lottery

A regional chamber of commerce holds a lottery for 50 businesses: 20 retail shops, 15 restaurants, 10 tech companies, and 5 consulting firms.

(a) What is the probability that the winner is a restaurant?
(b) What is the probability that the winner is either a tech company or a consulting firm?
(c) Two prizes are drawn without replacement. What is the probability that both winners are retail shops?

::: {.callout-tip collapse="true"}
## Solution

(a) $P(\text{restaurant}) = 15/50 = 0.30$.
(b) $P(\text{tech} \cup \text{consulting}) = 15/50 = 0.30$.
(c) $P(\text{both retail}) = \dfrac{20}{50} \cdot \dfrac{19}{49} = \dfrac{380}{2450} \approx 0.155$.
:::

### Exercise 3.3 ★★ — Inclusion-exclusion in a tax registry

In a city of 10,000 taxpayers: 3,500 have self-employment income (event $A$), 5,200 have salaried income (event $B$), and 1,400 have both.

(a) Compute $P(A \cup B)$.
(b) What is the probability of having neither source of income?
(c) Compute $P(A \cap \bar{B})$, the probability of self-employment income only.

*A full answer is given in the Instructor Edition.*

### Exercise 3.4 ★★ — Conditional probability from a two-way table

A human-resources department classifies 500 job applicants by education level and outcome of an aptitude test:

| | Pass | Fail | Total |
|---|---|---|---|
| Secondary education | 60 | 90 | 150 |
| University degree | 140 | 60 | 200 |
| Postgraduate | 120 | 30 | 150 |
| **Total** | **320** | **180** | **500** |

(a) Compute $P(\text{Pass})$.
(b) Compute $P(\text{Pass} \mid \text{University degree})$.
(c) Compute $P(\text{Postgraduate} \mid \text{Pass})$.
(d) Is the test result independent of education level? Justify with the appropriate check.

*A full answer is given in the Instructor Edition.*

### Exercise 3.5 ★★ — Law of total probability (insurance)

An insurance company classifies car-insurance policyholders by risk: low (60% of policies, claim probability 0.05), medium (30%, 0.15), high (10%, 0.40).

(a) What is the overall probability that a randomly selected policyholder files a claim?
(b) If 10,000 policies are issued, how many claims are expected?
(c) Premium is €400 per policy and the average claim costs €3,000. What is the expected profit per policy?

*A full answer is given in the Instructor Edition.*

### Exercise 3.6 ★★ — Bayes' theorem (medical test)

A rapid test for a rare disease (prevalence 1%) has sensitivity 98% and specificity 95%.

(a) Compute $P(+)$.
(b) Compute $P(D \mid +)$ using Bayes' theorem.
(c) Interpret the result — why is it surprisingly low?
(d) Recompute (b) if the prevalence were 10%. Compare.

*A full answer is given in the Instructor Edition.*

### Exercise 3.7 ★★★ — Monty Hall via Bayes

In the Monty Hall problem, let $C_i$ = "car is behind door $i$" and $H_3$ = "host opens door 3."

(a) State the prior probabilities $P(C_1), P(C_2), P(C_3)$.
(b) Compute $P(H_3 \mid C_1), P(H_3 \mid C_2), P(H_3 \mid C_3)$ assuming the host opens uniformly among permitted doors.
(c) Use Bayes' theorem to compute $P(C_1 \mid H_3)$ and $P(C_2 \mid H_3)$.
(d) Conclude: should the contestant switch?

*A full answer is given in the Instructor Edition.*

### Exercise 3.8 ★★★ — Multi-stage Bayes (warranty repairs)

A company sells Standard laptops (70% of sales) and Premium laptops (30%). Within the first year, 8% of Standard and 3% of Premium laptops need warranty repair. Among repairs, 60% of Standard and 40% of Premium repairs exceed €200.

(a) What is the overall probability that a laptop needs warranty repair?
(b) Given warranty repair, what is the probability the laptop is Standard?
(c) What is the probability that a laptop requires a warranty repair costing more than €200?
(d) Given an expensive repair (> €200), what is the probability the laptop is Premium?

*A full answer is given in the Instructor Edition.*
