---
title: "Random Variables"
---

> *Status: ported 2026-05-19. Reviewed by editor: pending.*

## Learning outcomes {.unnumbered}

By the end of this chapter the reader should be able to:

- Define a random variable as a function $X\colon \Omega \to \mathbb{R}$ and distinguish discrete from continuous random variables.
- Use the probability mass function $p(x)$ to describe a discrete random variable and verify that it is valid.
- Use the probability density function $f_X(x)$ to describe a continuous random variable and compute interval probabilities by integration.
- Construct the cumulative distribution function $F_X(x)$ in both the discrete and continuous cases and apply its monotonicity, right-continuity, and limit properties.
- Compute $\mathbb{E}[X]$ and $\mathbb{E}[g(X)]$ for discrete and continuous random variables and use the linearity of expectation.
- Compute $\operatorname{Var}(X)$ using the shortcut $\operatorname{Var}(X) = \mathbb{E}[X^2] - \mu^2$, and apply the rule $\operatorname{Var}(a + bX) = b^2 \sigma^2$.
- Standardise a random variable as $Z = (X - \mu)/\sigma$ and explain why $\mathbb{E}[Z] = 0$ and $\operatorname{Var}(Z) = 1$.
- State and apply Chebyshev's inequality to bound tail probabilities when the distribution is unknown.

## Motivating empirical question {.unnumbered}

> *A startup considering a new product launch faces three scenarios — moderate gain, large gain, or loss. What is its expected profit, and how risky is the venture?*

In Topic 3 we built the language of probability around abstract sample spaces. In practice, almost every business or economic question is numerical: how much profit, how many customers, what return? A **random variable** is the formal device that turns an abstract experiment into a real-valued quantity we can integrate, differentiate, and average. The running example throughout this chapter is the product-launch profit $X$, with three outcomes $-50$, $20$, $100$ (thousands of euros) and probabilities $0.20$, $0.50$, $0.30$ — concrete enough to compute, rich enough to illustrate every definition.

## 4.1 Random variables: from outcomes to numbers

In Topic 3 the elements of $\Omega$ could be anything — heads/tails, defective/non-defective, expansion/contraction. The full machinery of calculus and algebra only becomes available once outcomes are *numbers*.

::: {.callout-note}
## Definition: random variable

A **random variable** is a function $X\colon \Omega \to \mathbb{R}$ that assigns a real number $X(\omega)$ to every outcome $\omega \in \Omega$.
:::

By convention, random variables are written with uppercase letters ($X, Y, Z$) and their realisations with the matching lowercase letters ($x, y, z$). Thus $X$ is the (still uncertain) variable, while $x$ is a value it may take.

::: {.callout-note}
## Example: two coin tosses

Toss a fair coin twice. The sample space is $\Omega = \{HH, HT, TH, TT\}$. Define $X =$ number of heads. Then $X(TT) = 0$, $X(HT) = X(TH) = 1$, $X(HH) = 2$, and the distribution of $X$ is $P(X = 0) = 1/4$, $P(X = 1) = 1/2$, $P(X = 2) = 1/4$.
:::

### 4.1.1 Discrete vs continuous

::: {.callout-note}
## Definition: discrete random variable

A random variable $X$ is **discrete** if it takes only a finite or countably infinite number of values $x_1, x_2, \ldots$. The probability is concentrated on isolated points.
:::

::: {.callout-note}
## Definition: continuous random variable

A random variable $X$ is **continuous** if it takes any value in one or more intervals of $\mathbb{R}$. Its probability is described by a density, and the probability of any single point is zero.
:::

Economic examples help:

- **Discrete**: number of insurance claims per month; number of defective items in a shipment; credit rating category (AAA, AA, A, ...).
- **Continuous**: daily stock returns; household income; GDP growth rate; waiting time at a service counter.

## 4.2 Discrete distributions: the probability mass function

::: {.callout-note}
## Definition: probability mass function (PMF)

The **probability mass function** of a discrete random variable $X$ is
$p(x_i) = P(X = x_i)$ for $i = 1, 2, \ldots$
:::

A PMF is valid if and only if

$$
p(x_i) \geq 0 \text{ for every } i, \qquad \sum_{i} p(x_i) = 1.
$$

The collection of pairs $\{(x_i, p(x_i))\}$ is the **probability distribution** of $X$. For the two-coin-toss example, the PMF is $p(0) = 1/4$, $p(1) = 1/2$, $p(2) = 1/4$, and $1/4 + 1/2 + 1/4 = 1$ as required.

## 4.3 The cumulative distribution function

The cumulative distribution function (CDF) is defined for *every* random variable — discrete, continuous, or mixed — and is often the most convenient single object to work with.

::: {.callout-note}
## Definition: cumulative distribution function (CDF)

The **cumulative distribution function** of a random variable $X$ is
$F_X(x) = P(X \leq x)$ for $x \in \mathbb{R}$.
:::

The CDF is the probability of falling at or below $x$.

::: {.callout-note}
## Properties of the CDF

For any random variable $X$, the CDF $F_X$ satisfies:

1. $0 \leq F_X(x) \leq 1$ for all $x$.
2. **Monotone non-decreasing**: if $a < b$, then $F_X(a) \leq F_X(b)$.
3. **Limits**: $\lim_{x \to -\infty} F_X(x) = 0$ and $\lim_{x \to +\infty} F_X(x) = 1$.
4. **Right-continuous**: $\lim_{x \to a^+} F_X(x) = F_X(a)$.
5. For any $a < b$: $P(a < X \leq b) = F_X(b) - F_X(a)$.
:::

The shape of $F_X$ reveals the kind of variable:

- For a **discrete** $X$, $F_X$ is a **step function** — flat between possible values, jumping at each $x_i$ by exactly $p(x_i)$.
- For a **continuous** $X$, $F_X$ is a **smooth continuous curve** with no jumps.

::: {.callout-note}
## Example: CDF of the two-coin toss

For $X =$ number of heads in two tosses,
:::

$$
F_X(x) = \begin{cases}
0    & x < 0,\\
1/4  & 0 \leq x < 1,\\
3/4  & 1 \leq x < 2,\\
1    & x \geq 2.
\end{cases}
$$

Probabilities of intervals are then trivial: $P(0 < X \leq 1) = F_X(1) - F_X(0) = 3/4 - 1/4 = 1/2$.

## 4.4 Continuous distributions: the probability density function

For a continuous random variable we cannot assign positive probability to individual points — there are uncountably many of them, and the probabilities would not sum to one. Probabilities are described instead by a density.

::: {.callout-note}
## Definition: probability density function (PDF)
:::

A function $f_X(x)$ is a **probability density function** of a continuous random variable $X$ if

$$
f_X(x) \geq 0 \text{ for all } x, \qquad \int_{-\infty}^{\infty} f_X(x)\,dx = 1.
$$

Probabilities are computed as areas under the curve:

$$
P(a \leq X \leq b) = \int_a^b f_X(x)\,dx.
$$

The PDF and CDF of a continuous variable are linked by the fundamental theorem of calculus:

$$
F_X(x) = \int_{-\infty}^{x} f_X(t)\,dt, \qquad f_X(x) = F_X'(x).
$$

::: {.callout-warning}
## Common mistake: point probabilities of a continuous variable

For any continuous $X$ and any value $a$, $P(X = a) = \int_a^a f_X(x)\,dx = 0$. Therefore the inclusion or exclusion of endpoints is irrelevant:
$P(a \leq X \leq b) = P(a < X < b) = P(a \leq X < b) = P(a < X \leq b)$. (For discrete $X$, by contrast, $P(X = a)$ can be positive and the endpoint matters.)
:::

::: {.callout-note}
## Worked example: $f_X(x) = 2x$ on $[0, 1]$

Take $f_X(x) = 2x$ for $0 \leq x \leq 1$, zero elsewhere. **Validity**: $f_X \geq 0$ and $\int_0^1 2x\,dx = [x^2]_0^1 = 1$, both check. **Interval probability**: $P(0.3 \leq X \leq 0.7) = \int_{0.3}^{0.7} 2x\,dx = [x^2]_{0.3}^{0.7} = 0.49 - 0.09 = 0.40$. **CDF**: $F_X(x) = \int_0^x 2t\,dt = x^2$ for $0 \leq x \leq 1$, with $F_X(x) = 0$ for $x < 0$ and $F_X(x) = 1$ for $x > 1$. Check: $F_X(0.7) - F_X(0.3) = 0.49 - 0.09 = 0.40$, matching the direct integral.
:::

### 4.4.1 Discrete vs continuous: side-by-side

| Feature | Discrete | Continuous |
|---|---|---|
| Values | countable set | interval(s) of $\mathbb{R}$ |
| Probability described by | PMF $p(x_i) = P(X = x_i)$ | PDF $f_X(x) \geq 0$ |
| $P(X = a)$ | can be $> 0$ | always $= 0$ |
| $P(a \leq X \leq b)$ | $\sum_{a \leq x_i \leq b} p(x_i)$ | $\int_a^b f_X(x)\,dx$ |
| CDF shape | step function | smooth curve |

## 4.5 Expectation

The expected value of $X$ is the probabilistic analogue of the sample mean from Topic 1. It is the long-run average value of $X$ over infinitely many repetitions of the experiment — physically, the centre of gravity (balance point) of the probability distribution.

::: {.callout-note}
## Definition: expectation (mean)
:::

The **expected value** of a random variable $X$, denoted $\mathbb{E}[X]$ or $\mu$, is

$$
\mu = \mathbb{E}[X] = \sum_{i} x_i\, p(x_i) \quad \text{(discrete)}, \qquad
\mu = \mathbb{E}[X] = \int_{-\infty}^{\infty} x\, f_X(x)\,dx \quad \text{(continuous)}.
$$

::: {.callout-note}
## Example: two-coin toss

For $X =$ number of heads in two fair tosses,

$\mathbb{E}[X] = 0 \cdot \tfrac{1}{4} + 1 \cdot \tfrac{1}{2} + 2 \cdot \tfrac{1}{4} = 1$.

On average we expect one head in two tosses — consistent with intuition.
:::

::: {.callout-note}
## Example: $f_X(x) = 2x$

For the density $f_X(x) = 2x$ on $[0, 1]$,

$\mathbb{E}[X] = \int_0^1 x \cdot 2x\,dx = \int_0^1 2x^2\,dx = \tfrac{2}{3}$.

The mean sits above $1/2$ because the density places more mass near $x = 1$.
:::

### 4.5.1 Expectation of a function: $\mathbb{E}[g(X)]$

For any (measurable) function $g$, the expected value of $g(X)$ is computed by weighting $g(x)$ by the probability mass or density at $x$:

$$
\mathbb{E}[g(X)] = \sum_i g(x_i)\, p(x_i) \quad \text{(discrete)}, \qquad
\mathbb{E}[g(X)] = \int g(x) f_X(x)\,dx \quad \text{(continuous)}.
$$

The most-used case is $g(x) = x^2$, which gives the **second moment** $\mathbb{E}[X^2]$ — the building block of variance below.

### 4.5.2 Linearity of expectation

For any constants $a, b$ and any random variable $X$,

$$
\mathbb{E}[a + bX] = a + b\,\mathbb{E}[X].
$$

This identity is fundamental: it requires no assumption on the distribution of $X$. It also extends to sums of random variables, $\mathbb{E}[X + Y] = \mathbb{E}[X] + \mathbb{E}[Y]$, even when $X$ and $Y$ are dependent (a fact we will exploit in Topic 5 for sums of Bernoulli indicators).

::: {.callout-note}
## Example: linear transformation in the bookshop

Suppose the daily number of textbooks sold $X$ has $\mathbb{E}[X] = 1.95$. Each book is sold for 25 euros, with daily fixed cost 30 euros, so daily profit is $Y = 25 X - 30$. Then $\mathbb{E}[Y] = 25 \cdot 1.95 - 30 = 18.75$ euros.
:::

## 4.6 Variance and standard deviation

The expected value locates the centre of the distribution; the variance measures how far values typically fall from that centre.

::: {.callout-note}
## Definition: variance and standard deviation
:::

The **variance** of $X$ is

$$
\sigma^2 = \operatorname{Var}(X) = \mathbb{E}\!\left[(X - \mu)^2\right] = \mathbb{E}[X^2] - \mu^2,
$$

and the **standard deviation** is $\sigma = \sqrt{\sigma^2}$, expressed in the same units as $X$.

Explicitly:

$$
\sigma^2 = \sum_i (x_i - \mu)^2\, p(x_i) = \sum_i x_i^2\, p(x_i) - \mu^2 \quad \text{(discrete)},
$$

$$
\sigma^2 = \int_{-\infty}^{\infty} (x - \mu)^2 f_X(x)\,dx = \int_{-\infty}^{\infty} x^2 f_X(x)\,dx - \mu^2 \quad \text{(continuous)}.
$$

::: {.callout-warning}
## Common mistake: the wrong sign in the shortcut

The shortcut formula is $\operatorname{Var}(X) = \mathbb{E}[X^2] - (\mathbb{E}[X])^2$ — the mean of the *squares* minus the *square* of the mean. Swapping these terms gives a negative number and a tell-tale sign that something has gone wrong. Use $S^2$ in Topic 1 as a memory anchor: same structure, sample analogues.
:::

::: {.callout-note}
## Properties of variance

For any random variable $X$ with $\operatorname{Var}(X) = \sigma^2$ and any constants $a, b$:

1. $\operatorname{Var}(X) \geq 0$, with equality iff $X$ is constant almost surely.
2. $\operatorname{Var}(a) = 0$.
3. $\operatorname{Var}(a + bX) = b^2 \sigma^2$. Shifts do not change variance; scaling by $b$ multiplies variance by $b^2$.
4. Consequently, $\operatorname{SD}(a + bX) = |b|\,\sigma$.
:::

::: {.callout-note}
## Example: defective tyres

A tyre factory inspects batches of three. Let $X$ count the defective tyres in a batch, with PMF $p(0) = 0.70$, $p(1) = 0.20$, $p(2) = 0.08$, $p(3) = 0.02$. Then $\mathbb{E}[X] = 0.42$, $\mathbb{E}[X^2] = 0.70$, so $\operatorname{Var}(X) = 0.70 - 0.42^2 = 0.5236$ and $\sigma = \sqrt{0.5236} \approx 0.724$.
:::

::: {.callout-note}
## Example: variance for $f_X(x) = 2x$

We already have $\mathbb{E}[X] = 2/3$ for the density $f_X(x) = 2x$. Compute $\mathbb{E}[X^2] = \int_0^1 x^2 \cdot 2x\,dx = \int_0^1 2x^3\,dx = 1/2$. Hence $\operatorname{Var}(X) = 1/2 - (2/3)^2 = 1/2 - 4/9 = 1/18 \approx 0.0556$, and $\sigma = 1/(3\sqrt{2}) \approx 0.236$.
:::

### 4.6.1 Notation bridge: descriptive vs probabilistic

In Topic 1 we computed *sample* statistics from observed data. We are now working with *population* parameters defined by a probability model. The correspondence is the standard Latin–Greek bridge:

| Concept | Sample (Topic 1) | Population (Topic 4) |
|---|---|---|
| Mean | $\bar{x}$ | $\mu = \mathbb{E}[X]$ |
| Variance | $S^2$ | $\sigma^2 = \operatorname{Var}(X)$ |
| Standard deviation | $S$ | $\sigma$ |
| Correlation | $r$ | $\rho$ |

In all of inferential statistics, the central question is how well a sample statistic estimates the corresponding population parameter — a story for TC2 / Econometrics I, not for this book.

## 4.7 Standardisation

A useful operation, used everywhere in Topic 5, is to centre and scale a random variable so that it has mean zero and variance one.

::: {.callout-note}
## Definition: standardised variable
:::

The **standardised** version of a random variable $X$ with mean $\mu$ and standard deviation $\sigma > 0$ is

$$
Z = \frac{X - \mu}{\sigma}.
$$

By the linearity of expectation and the scaling rule for variance,

$$
\mathbb{E}[Z] = \frac{\mathbb{E}[X] - \mu}{\sigma} = 0, \qquad \operatorname{Var}(Z) = \frac{1}{\sigma^2}\operatorname{Var}(X) = 1.
$$

A value of $Z$ measures how many standard deviations $X$ is away from its mean. Topic 5 introduces the **standard normal** distribution (a particular continuous $Z$ with mean $0$, variance $1$, and bell-shaped density) and the table of its CDF $\Phi(z)$ — but standardisation itself is a purely mechanical operation, available for *any* variable with finite variance.

## 4.8 Chebyshev's inequality

How likely is it that $X$ lies far from its mean? Without knowing the shape of the distribution, we can still give a universal answer in terms of the standard deviation.

::: {.callout-note}
## Chebyshev's inequality
:::

For any random variable $X$ with mean $\mu$ and finite variance $\sigma^2$, and for any $k > 0$,

$$
P(|X - \mu| \geq k\sigma) \leq \frac{1}{k^2},
$$

equivalently,

$$
P(\mu - k\sigma \leq X \leq \mu + k\sigma) \geq 1 - \frac{1}{k^2}.
$$

In words: regardless of distribution shape, at least a fraction $1 - 1/k^2$ of the probability lies within $k$ standard deviations of the mean.

::: {.callout-note}
## Example: mutual-fund returns

A fund has expected return $\mu = 8\%$ and $\sigma = 6\%$. With no assumption about the distribution: $P(-4\% \leq X \leq 20\%) \geq 1 - 1/4 = 0.75$ (taking $k = 2$); $P(-10\% \leq X \leq 26\%) \geq 1 - 1/9 \approx 0.889$ (taking $k = 3$). If we want at least 95% probability, we need $1 - 1/k^2 \geq 0.95$, i.e. $k \geq \sqrt{20} \approx 4.47$, giving the wide interval $8\% \pm 26.8\%$.
:::

::: {.callout-warning}
## Chebyshev is universal but loose

Chebyshev makes *no* assumption about the distribution. The price is that the bound is typically very conservative. Under normality (Topic 5), $P(|X - \mu| \leq 2\sigma) \approx 0.954$, far above the Chebyshev lower bound of $0.75$. Use Chebyshev when the distribution is unknown; use distribution-specific tables when it is known.
:::

<!-- TODO: Markov's inequality is not stated in the Topic 4 lecture notes (only Chebyshev). If the editor wants it included as a precursor to Chebyshev (Markov for non-negative variables: $P(X \geq a) \leq \mathbb{E}[X]/a$), insert here. -->

## 4.9 R Lab — Discrete and continuous random variables

This lab walks through the running product-launch example (discrete) and the density $f_X(x) = 2x$ (continuous): plotting the PMF/PDF and CDF, and computing the mean and variance directly from the definitions.

```{r ch04-setup}
#| message: false
#| warning: false
set.seed(2026)
# No special packages required: base R only.
```

### 4.9.1 Discrete: product-launch profit

```{r ch04-discrete-pmf}
outcomes <- c(-50, 20, 100)            # profit in thousands of euros
probs    <- c(0.20, 0.50, 0.30)

pmf <- data.frame(x = outcomes, p = probs)
knitr::kable(pmf, caption = "PMF of product-launch profit (k EUR)")

barplot(probs, names.arg = outcomes, col = "steelblue",
        xlab = "Profit (k EUR)", ylab = "P(X = x)",
        main = "PMF: product-launch profit", las = 1, ylim = c(0, 0.6))
```

```{r ch04-discrete-mean-var}
EX   <- sum(outcomes * probs)
EX2  <- sum(outcomes^2 * probs)
VarX <- EX2 - EX^2
SDX  <- sqrt(VarX)

cat("E[X]  =", EX,           "k EUR\n")
cat("E[X^2]=", EX2,           "\n")
cat("Var(X)=", VarX,          "\n")
cat("SD(X) =", round(SDX, 2), "k EUR\n")
```

The expected profit is 30k EUR but the standard deviation is roughly 48k EUR — larger than the mean itself, which signals a risky venture.

```{r ch04-discrete-cdf}
cum_probs <- cumsum(probs)
cdf <- data.frame(x = outcomes, F_x = cum_probs)
knitr::kable(cdf, caption = "CDF values (jumps of the step function)")

# Hand-built step plot
x_plot <- c(-80, outcomes[1], outcomes[1], outcomes[2],
            outcomes[2], outcomes[3], outcomes[3], 130)
y_plot <- c(0, 0, cum_probs[1], cum_probs[1],
            cum_probs[2], cum_probs[2], cum_probs[3], cum_probs[3])
plot(x_plot, y_plot, type = "l", col = "steelblue", lwd = 2,
     xlab = "x (k EUR)", ylab = "F(x) = P(X <= x)",
     main = "CDF: product-launch profit", las = 1, ylim = c(0, 1))
points(outcomes, cum_probs, pch = 19, col = "steelblue", cex = 1.2)
```

We can read off $P(X \leq 20) = 0.70$: a 70% chance that profit does **not** exceed 20k EUR.

### 4.9.2 Continuous: $f_X(x) = 2x$ on $[0, 1]$

```{r ch04-continuous-pdf}
f <- function(x) ifelse(x >= 0 & x <= 1, 2 * x, 0)

total <- integrate(f, 0, 1)$value
cat("Integral of f from 0 to 1:", total, "(should be 1)\n")

curve(f, from = -0.2, to = 1.3, n = 300, col = "steelblue", lwd = 2,
      xlab = "x", ylab = "f(x)", main = "PDF: f(x) = 2x", las = 1)
x_shade <- seq(0.3, 0.7, length.out = 100)
polygon(c(0.3, x_shade, 0.7), c(0, f(x_shade), 0),
        col = rgb(0.27, 0.51, 0.71, 0.3), border = NA)
```

```{r ch04-continuous-probs}
p_mid   <- integrate(f, 0.3, 0.7)$value
EX_c    <- integrate(function(x) x * f(x),   0, 1)$value
EX2_c   <- integrate(function(x) x^2 * f(x), 0, 1)$value
VarX_c  <- EX2_c - EX_c^2

cat("P(0.3 <= X <= 0.7) =", round(p_mid, 4),  "\n")
cat("E[X]               =", round(EX_c, 4),   " (theory: 2/3)\n")
cat("Var(X)             =", round(VarX_c, 4), " (theory: 1/18)\n")
```

```{r ch04-continuous-cdf}
F_cdf <- function(x) ifelse(x < 0, 0, ifelse(x > 1, 1, x^2))

curve(F_cdf, from = -0.3, to = 1.3, n = 300, col = "steelblue", lwd = 2,
      xlab = "x", ylab = "F(x)", main = "CDF: F(x) = x^2 on [0, 1]", las = 1)
abline(h = c(0, 1), col = "grey80", lty = 3)
```

By the fundamental theorem of calculus, $F_X(0.7) - F_X(0.3) = 0.49 - 0.09 = 0.40$ matches the direct integral.

### 4.9.3 Chebyshev in practice

```{r ch04-chebyshev}
# Discrete product-launch variable: use Chebyshev to bound P(|X - mu| < k*sigma)
mu_X <- EX
sd_X <- SDX
k_vals <- c(1, 2, 3, 4)
lower_bound <- 1 - 1/k_vals^2
data.frame(k = k_vals,
           interval_lo = round(mu_X - k_vals * sd_X, 1),
           interval_hi = round(mu_X + k_vals * sd_X, 1),
           Cheb_lower  = round(lower_bound, 3))
```

For our product launch, Chebyshev guarantees that with probability at least 0.75 the profit lies within $\mu \pm 2\sigma$, i.e. between $-65.6$ and $125.6$ thousand euros — a wide and very conservative band, but distribution-free.

## Self-check {.unnumbered}

::: {.callout-tip collapse="true"}
## Q1. Defining property of a PMF

A function $p(x)$ is a valid probability mass function for a discrete random variable if and only if:

- A. $p(x) > 0$ for every $x$ in the support.
- B. $p(x) \geq 0$ for every $x$ and $\sum_x p(x) = 1$.
- C. $\int p(x)\,dx = 1$.
- D. $p(x) = P(X \leq x)$ for every $x$.

**Answer: B.** Non-negativity *plus* probabilities summing to one. Strict positivity in A is too strong (zero is allowed); C is the PDF condition; D is the CDF, not the PMF.
:::

::: {.callout-tip collapse="true"}
## Q2. CDF of a discrete random variable

The cumulative distribution function $F_X(x) = P(X \leq x)$ of a discrete random variable is:

- A. A non-decreasing step function with jumps equal to $p(x)$ at each possible value.
- B. A continuous straight line from $0$ to $1$.
- C. A function that decreases from $1$ to $0$.
- D. Identical to the PMF.

**Answer: A.** The CDF accumulates probability and jumps by exactly $p(x_i)$ at each support point. It is flat between jumps and is right-continuous at each jump.
:::

::: {.callout-tip collapse="true"}
## Q3. Point probability for a continuous variable

For a continuous random variable $X$ and any specific value $x_0$, $P(X = x_0)$ equals:

- A. $f_X(x_0)$.
- B. $F_X(x_0)$.
- C. Exactly zero — only intervals have positive probability.
- D. $1/n$, where $n$ is the sample size.

**Answer: C.** Because $\int_{x_0}^{x_0} f_X = 0$. The density value $f_X(x_0)$ is *not* a probability; it has units of probability per unit of $x$.
:::

::: {.callout-tip collapse="true"}
## Q4. Probability from the PDF $f_X(x) = 2x$

If $f_X(x) = 2x$ on $[0, 1]$ (and zero elsewhere), then $P(X \leq 0.5)$ equals:

- A. $f_X(0.5) = 1$.
- B. $F_X(0.5) = 0.5^2 = 0.25$.
- C. $0.5$ (half of the interval).
- D. $2 \cdot 0.5 = 1$.

**Answer: B.** $P(X \leq 0.5) = \int_0^{0.5} 2x\,dx = 0.25$. Most of the mass lies above $0.5$ because the density is increasing.
:::

::: {.callout-tip collapse="true"}
## Q5. The shortcut formula for variance

The variance of a random variable can always be written as:

- A. $\operatorname{Var}(X) = \mathbb{E}[X^2] - (\mathbb{E}[X])^2$.
- B. $\operatorname{Var}(X) = \mathbb{E}[X] - (\mathbb{E}[X])^2$.
- C. $\operatorname{Var}(X) = (\mathbb{E}[X^2])^2 - \mathbb{E}[X]$.
- D. $\operatorname{Var}(X) = (\mathbb{E}[X])^2 - \mathbb{E}[X^2]$.

**Answer: A.** The mean of the squares minus the square of the mean. Option D has the sign reversed and would be non-positive.
:::

::: {.callout-tip collapse="true"}
## Q6. Linear transformation

If $\mathbb{E}[X] = 50$ and $\operatorname{Var}(X) = 4$, and $Y = 200 + 10 X$, then:

- A. $\mathbb{E}[Y] = 700$ and $\operatorname{Var}(Y) = 4$.
- B. $\mathbb{E}[Y] = 700$ and $\operatorname{Var}(Y) = 400$.
- C. $\mathbb{E}[Y] = 250$ and $\operatorname{Var}(Y) = 40$.
- D. $\mathbb{E}[Y] = 700$ and $\operatorname{Var}(Y) = 40$.

**Answer: B.** $\mathbb{E}[Y] = 200 + 10 \cdot 50 = 700$ by linearity; $\operatorname{Var}(Y) = 10^2 \cdot 4 = 400$ because the constant 200 does not affect the variance and the multiplier 10 enters squared.
:::

::: {.callout-tip collapse="true"}
## Q7. Standardisation

If $Z = (X - \mu)/\sigma$ is the standardised version of $X$, then:

- A. $\mathbb{E}[Z] = \mu$ and $\operatorname{Var}(Z) = \sigma^2$.
- B. $\mathbb{E}[Z] = 0$ and $\operatorname{Var}(Z) = 1$.
- C. $Z$ is always normally distributed.
- D. $\mathbb{E}[Z] = 1$ and $\operatorname{Var}(Z) = 0$.

**Answer: B.** Linearity of expectation gives $\mathbb{E}[Z] = 0$, and the variance scaling rule gives $\operatorname{Var}(Z) = 1$. Option C is a common confusion: standardisation centres and scales but does *not* change the shape of the distribution.
:::

::: {.callout-tip collapse="true"}
## Q8. Chebyshev's inequality

A variable has $\mu = 500$, $\sigma = 40$. By Chebyshev, the minimum probability that $X$ falls in the interval $(380, 620)$ is at least:

- A. $1/3$.
- B. $1/2$.
- C. $8/9$.
- D. $1$.

**Answer: C.** The interval is $\mu \pm 3\sigma$, so $k = 3$ and the lower bound is $1 - 1/9 = 8/9 \approx 0.889$.
:::

## Exercises {.unnumbered}

### Exercise 4.1 ★ — Validating a PMF and computing $\mathbb{E}[X]$

A small consulting firm receives $X$ client enquiries per day. The proposed PMF is

| $x$ | 0 | 1 | 2 | 3 | 4 |
|---|---|---|---|---|---|
| $P(X = x)$ | 0.10 | 0.25 | 0.35 | 0.20 | $k$ |

(a) Find $k$ so that this is a valid PMF.
(b) Compute $\mathbb{E}[X]$.
(c) Compute $P(X \geq 2)$.

::: {.callout-tip collapse="true"}
## Solution

(a) Probabilities must sum to one: $0.10 + 0.25 + 0.35 + 0.20 + k = 1$, so $k = 0.10$.

(b) $\mathbb{E}[X] = 0(0.10) + 1(0.25) + 2(0.35) + 3(0.20) + 4(0.10) = 0.25 + 0.70 + 0.60 + 0.40 = 1.95$.

(c) $P(X \geq 2) = 0.35 + 0.20 + 0.10 = 0.65$.
:::

### Exercise 4.2 ★ — Variance via the shortcut formula

Using the PMF from Exercise 4.1 (with $k = 0.10$):

(a) Compute $\mathbb{E}[X^2]$.
(b) Compute $\operatorname{Var}(X) = \mathbb{E}[X^2] - (\mathbb{E}[X])^2$.
(c) Report $\sigma$.

::: {.callout-tip collapse="true"}
## Solution

(a) $\mathbb{E}[X^2] = 0 + 0.25 + 4(0.35) + 9(0.20) + 16(0.10) = 0.25 + 1.40 + 1.80 + 1.60 = 5.05$.

(b) $\operatorname{Var}(X) = 5.05 - 1.95^2 = 5.05 - 3.8025 = 1.2475$.

(c) $\sigma = \sqrt{1.2475} \approx 1.117$.
:::

### Exercise 4.3 ★★ — Recovering the PMF from a step CDF

A discrete random variable $X$ has CDF

$$
F_X(x) = \begin{cases}
0    & x < 1,\\
0.15 & 1 \leq x < 3,\\
0.40 & 3 \leq x < 5,\\
0.75 & 5 \leq x < 7,\\
1    & x \geq 7.
\end{cases}
$$

(a) Recover the PMF from the jumps of $F_X$.
(b) Compute $P(3 < X \leq 7)$.
(c) Compute $\mathbb{E}[X]$ and $\operatorname{Var}(X)$.

### Exercise 4.4 ★★ — Continuous PDF: validity, probabilities, CDF

The time $X$ (in hours) a customer spends in a shopping centre has PDF $f_X(x) = c\,x(4 - x)$ on $[0, 4]$, zero elsewhere.

(a) Find $c$.
(b) Compute $P(1 \leq X \leq 3)$.
(c) Find the CDF $F_X(x)$ on $[0, 4]$.
(d) Find the median by solving $F_X(m) = 0.5$.

### Exercise 4.5 ★★ — Linear transformation of a random variable

The monthly electricity bill $X$ (in euros) of a household has $\mathbb{E}[X] = 85$ and $\operatorname{Var}(X) = 225$. A government subsidy transforms the bill to $Y = 0.80\,X - 10$.

(a) Compute $\mathbb{E}[Y]$.
(b) Compute $\operatorname{Var}(Y)$ and $\sigma_Y$.
(c) What is the average monthly saving, $\mathbb{E}[X - Y]$?
(d) Does the subsidy reduce the variability of bills? By how much?

### Exercise 4.6 ★★ — Mean and variance of a continuous variable

The weekly profit $X$ (in thousands of euros) of a small online retailer has

$$
f_X(x) = \begin{cases} \tfrac{1}{18}(6 - x) & 0 \leq x \leq 6,\\ 0 & \text{otherwise.}\end{cases}
$$

(a) Verify that $f_X$ is a valid PDF.
(b) Compute $\mathbb{E}[X]$.
(c) Compute $\mathbb{E}[X^2]$ and $\operatorname{Var}(X)$.
(d) If the retailer faces a fixed weekly cost of 1,500 EUR, what is the expected net profit $\mathbb{E}[X - 1.5]$ (in thousands of euros)?

### Exercise 4.7 ★★★ — Chebyshev for an e-commerce warehouse

The daily number of online orders $X$ at a warehouse has $\mathbb{E}[X] = 500$ and $\sigma = 40$.

(a) Using Chebyshev, find the minimum probability that $X$ falls between 380 and 620.
(b) The warehouse has capacity for 650 orders per day. Use Chebyshev to give an upper bound on the probability of exceeding capacity.
(c) Find the value of $k$ for which Chebyshev guarantees probability at least 0.95 of being within $\mu \pm k\sigma$, and translate this back into an interval for $X$.
(d) Compare your answer in (a) with the value that would be obtained under normality (Topic 5: $P(\mu - 2\sigma \leq X \leq \mu + 2\sigma) \approx 0.954$). What does the gap tell you about the cost of distribution-free bounds?

### Exercise 4.8 ★★★ — A symmetric continuous distribution

A financial analyst models the monthly return $R$ (in %) of a stock as a continuous variable on $[-10, 10]$ with PDF $f_R(r) = \tfrac{3}{4000}(100 - r^2)$.

(a) Verify that $f_R$ is a valid PDF.
(b) Compute $P(-5 \leq R \leq 5)$.
(c) Use symmetry to show $\mathbb{E}[R] = 0$.
(d) Compute $\operatorname{Var}(R)$ and $\sigma_R$.
(e) Apply Chebyshev's inequality to bound $P(-10 \leq R \leq 10)$ and compare with the exact answer.

<!--
Cross-references:
- Builds on Topic 3 (probability foundations, sample spaces, $P(A)$).
- Forward reference: Topic 5 introduces specific named distributions (Bernoulli, binomial, Poisson, uniform, exponential, normal) and the standard normal table for $\Phi(z)$.
- The bivariate material in the source lecture notes (joint distributions, marginals, covariance, correlation, independence, portfolio variance) is rich and present in the R Lab, but lies outside the scope of this chapter as listed by the editor. It can be folded into Topic 5 or a stand-alone chapter on bivariate / joint distributions if desired. TODO for editor.
-->
